### 2023-05-04 README.Rmd
Create README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# FiringSeriation

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2023-05-17 analysis/supplementary-materials/01-data.R
Pre-process XRD data

@@ -1,0 +1,136 @@
+# HELPERS ======================================================================
+#' Get Sample Code from File Names
+#'
+#' @param file A [`character`] vector of path names.
+#' @return A [`character`] vector.
+#' @noRd
+get_bdx <- function(file) {
+  regmatches(x = file, m = regexpr(pattern = "BDX[0-9]{5}", text = file))
+}
+
+# READ XRD =====================================================================
+## List all diffractograms (Brucker RAW)
+xrd_files <- here::here("analysis/data/raw_data") |>
+  list.files(pattern = ".raw", full.names = TRUE)
+
+## List all samples
+xrd_samples <- get_bdx(xrd_files)
+
+## Read files
+xrd_data <- lapply(
+  X = xrd_files,
+  FUN = function(x) {
+    diffractogram <- rxylib::read_xyData(x, verbose = FALSE, metaData = FALSE)
+    xy <- diffractogram$dataset[[1]]$data_block
+    colnames(xy) <- c("theta", get_bdx(x))
+    as.data.frame(xy)
+  }
+)
+
+## Check that all diffractograms share the same scale (2 theta)
+theta_scale <- xrd_data[[1]]$theta
+for (i in 2:length(xrd_data)) {
+  if (isFALSE(all.equal(theta_scale, xrd_data[[i]])))
+     warning(sprintf("The theta positions of sample %d is different.", i))
+}
+
+## Visual inspection
+plot(
+  x = NA, y = NA,
+  xlab = "", ylab = "",
+  xlim = c(3, 70),
+  ylim = c(0, 120000)
+)
+for (i in xrd_data) {
+  lines(x = i[[1]], y = i[[2]])
+}
+
+# PRE-PROCESS ==================================================================
+## Penalized likelihood smoothing
+lambda <- seq(from = 1, to = 8, length.out = 40)
+lambda <- 10^lambda
+
+xrd_clean <- vector(mode = "list", length = length(xrd_data))
+for (i in seq_along(xrd_data)) {
+  ## Save plot for visual inspection
+  sprintf("analysis/figures/process_%s.png", xrd_samples[i]) |>
+    here::here() |>
+    png(width = 1024, height = 768, res = 100)
+
+  selected <- xrd_data[[i]]
+  names(selected) <- c("x", "y")
+
+  ## Subset from 5° to 55° (2theta)
+  selected <- alkahest::signal_select(selected, from = 3.5, to = 70)
[…]

### 2023-05-17 analysis/supplementary-materials/02-abstract.R
Add graphical abstract

@@ -1,0 +1,87 @@
+## GRAPHICAL ABSTRACT
+
+## Install rayshader (if needed)
+if (!requireNamespace("rayshader", quietly = TRUE))
+  install.packages("rayshader")
+
+## Read XRD data
+## (see analysis/supplementary-materials/01-data.R)
+xrd_counts <- here::here("analysis/data/derived_data/xrd_clean.csv") |>
+  read.table(sep = ",", dec = ".", header = TRUE)
+
+## Remove the first column (2 theta positions)
+theta_scale  <- xrd_counts[, 1]
+xrd_counts <- xrd_counts[, -1]
+
+## Transpose (one sample per line)
+xrd_counts <- xrd_counts |> t() |> as.data.frame()
+colnames(xrd_counts) <- paste0("T", seq_len(ncol(xrd_counts)))
+
+## Read archaeological data
+corpus <- here::here("analysis/data/raw_data/corpus.csv") |>
+  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8")
+
+## Mas de Moreno: ceramics + unfired sherds
+# all(corpus$sample == rownames(xrd_counts))
+index_moreno <- corpus$site == "Mas de Moreno"
+
+## Correspondance analysis
+ca_moreno <- dimensio::ca(xrd_counts[index_moreno, ])
+
+## Get CA order
+ca_coords <- dimensio::get_coordinates(ca_moreno)
+ca_index <- order(ca_coords[, 1], decreasing = TRUE)
+
+## Reorder matrix
+xrd_counts2 <- xrd_counts[ca_index, ]
+colnames(xrd_counts2) <- as.character(theta_scale)
+xrd_counts2$sample <- factor(rownames(xrd_counts2), levels = rownames(xrd_counts2))
+
+gg_ray <- xrd_counts2 |>
+  tidyr::gather(key = "theta", value = "value", -sample) |>
+  dplyr::filter(theta >= 25 & theta <= 32) |>
+  dplyr::mutate(
+    theta = as.numeric(theta),
+    sample2 = as.numeric(sample),
+    # value = sqrt(value),
+    value = value / 1000
+  ) |>
+  ggplot2::ggplot() +
+  ggplot2::geom_tile(ggplot2::aes(x = theta, y = sample2, fill = value)) +
+  ggplot2::scale_x_continuous(breaks = seq(25, 32, 1), expand = c(0, 0)) +
[…]

### 2023-05-18 analysis/paper/paper.qmd
Draft introduction

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/elsevier-harvard.csl"
abstract: |
  Text of abstract
keywords: |
  XRD; ceramic firing; correspondence analysis; matrix seriation
highlights: |
  These are the highlights. 
---

```{r}
#| label: setup
options(knitr.kable.NA = "")
```

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

```{r}
#| label: graphical-abstract
#| fig-align: center
#| out.width: "4in"
knitr::include_graphics(here::here("analysis/figures/abstract.png"))
```

\newpage

# License {-}

This work is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).

# Change record {-}

| **Version** | **Date**   | **Description**            | **DOI** |
|------------:|-----------:|:---------------------------|:--------|
| 0.0.0.9000  |  |  |  |

\newpage

<!-- The actual document text starts here: -->

# INTRODUCTION

Firing is a "strategic step" [@lemonnier1983a] of the ceramic production process, especially as it involves an economic risk: that of losing all or part of the load to be fired and as much human labour time. The study of the firing of archaeological ceramics is thus a key factor in the interpretation of ancient techniques and the related social practices. To this end, important methodological research has been carried out since the pioneering work of @shepard1936. This led to the wide use of the so-called *Equivalent Firing Temperatures* (EFT) as a way to describe the firing of archaeological ceramics [@tite1969; @tite1969a]. The EFT is a point estimator tied to standard experimental conditions: "[EFT] is that temperature maintained for one hour which would produce the observed mineralogy or microstructure" [@tite1995].

The EFT is certainly practical, but it raises several methodological and conceptual problems. A full review is beyond the scope of this study, but two points can be made. On the one hand, EFT creates unnecessary ambiguity about the real firing temperature (assuming that the firing temperatures have archaeological significance). On the other hand, @gosselain1992 and @livingstonesmith2001 have demonstrated the limits of this framework, which may lead to anachronistic reasoning [@frerebeau2018].

This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data, which aims to provide an alternative to the EFT framework. This method is based on the principle of parsimony with two main objectives: (1) to avoid the use of external references (whether experimental or ethnographic data); (2) to build the simplest possible model by limiting the number of hypotheses involved.

XRD analysis from archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing. Quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content. The estimation of quantitative mineralogical compositions, by the RIR method [@hubbard1988] or the Rietvelt refinement [@rietveld1969], is a very difficult, high skilled exercise, assuming a thorough knowledge of the material (and expensive databases).

In this study, we show that it is possible to successfully use correspondence analysis of X-ray diffractograms to assess the firing of ceramic materials. The properties of correspondence analysis and its links with matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers. Thus, calibration data sets can be produced on the only basis of archaeological evidence to assess ceramic firing practices across a cultural area, avoiding the use of laboratory references.

As an illustration of this method, we examine XRD data of Iron Age ceramics from an Iberian workshop. The Mas de Moreno workshop is located in the province of Teruel (Aragon, Spain), in the municipality of Foz-Calanda. The activity of the workshop is established between 225-200 BC and 40-30 AD. A first archaeological synthesis of the workshop has highlighted three successive phases of activity, characterised by successive changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012]. Several kilns have been excavated, which were discontinued and destroyed around 50-40 BC [@gorgues2012]. The use of an amphora kiln, of much larger dimensions than the previous kilns, as well as the production of Roman amphora and the presence of Latin epigraphy, illustrate a major reorganisation of the workshop in the middle of the 1^st^ century. The workshop's activity ceased during the following decade, mostly as a result of the territorial reorganisation under Roman influence and the distance from the trade routes [@gorgues2012]. This workshop thus constitutes a context well suited to investigate the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.

# METHODS

# RESULTS

# DISCUSSION

# CONCLUSION

# ACKNOWLEDGEMENTS

This study has received a State financial support managed by the French Agence Nationale de la Recherche under the program *Investissements d'avenir* (ANR-10-LABX-52).

The author is grateful to Alexis Gorgues (UMR 5607 Institut Ausonius) and José Antonio Benavente (Consorcio Patrimonio Ibérico de Aragón) for providing access to archaeological material.

# DATA AVAILABILITY STATEMENT

The data that support the findings of this study are openly available in [@frerebeau2015h] and [@frerebeau2015i].

Code for data cleaning and analysis is provided as compendium in @frerebeau2023a.

\newpage

# REFERENCES

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

# Colophon {-}

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
sessionInfo()
```

### 2023-05-18 analysis/paper/paper.qmd
Draft Methods section

@@
 
-XRD analysis from archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing.
+XRD analysis of archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing.
 Quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content.
@@
 
-In this study, we show that it is possible to successfully use correspondence analysis of X-ray diffractograms to assess the firing of ceramic materials.
+In this study, we show that it is possible to successfully use correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials.
 The properties of correspondence analysis and its links with matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers.
@@
 The Mas de Moreno workshop is located in the province of Teruel (Aragon, Spain), in the municipality of Foz-Calanda.
-The activity of the workshop is established between 225-200 BC and 40-30 AD.
+The activity of the workshop spans between 225-200 BC and 40-30 AD.
 A first archaeological synthesis of the workshop has highlighted three successive phases of activity, characterised by successive changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012].
@@
 # METHODS
+## Archaeological samples
+
+Sampling was carried out in such a way that each sample was taken from a single artifact.
+Particular attention was paid to the risk of contamination by systematically discarding sherds that had been glued back together or that had undergone any chemical treatment.
+
+```{r}
+#| label: archaeological-data
+## Read archaeological data
+corpus <- here::here("analysis/data/raw_data/corpus.csv") |>
+  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8")
+```
+
+`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop were sampled to represent the entire period of the workshop's activity (@tbl-corpus).
+The ceramic material of the Mas de Moreno consists only of products known as Iberian painted/fine wares.
+It refers to a wheel thrown production, with a pink uniform colour and a dark red painted decor, with no temper visible to the naked eye and considered to be fired at high temperatures in an oxidizing atmosphere [@tarradell1980; @mata1992; @collconesa2000].
+
+```{r}
+#| label: tbl-corpus
+#| tbl-cap: "Ceramic and clay samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. LOI: loss on ignition [see @frerebeau2020a]."
+knitr::kable(
+  x = corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration"))],
+  col.names = c("Sample", "Material", "Site", "Stratigraphic Unit", "Phase", "Type", "Part", "LOI (%)")
+)
+```
+
+Although Iberian fine wares present a significant technological uniformity [@tarradell1980], the sampling was conducted in such a way that all types of pottery produced in the workshop were included in the study (non-Iberian types produced at the end of the workshop's activity period were also sampled).
+
+A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017].
+Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (@tbl-corpus).
+
+Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (@tbl-corpus).
+These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dates between the 8^th^ and the 4^th^ century BC).
+Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study.
+
+## Powder X-ray diffraction
+The mineralogical composition of the samples was determined by X-ray powder diffraction.
+The outer surfaces of all the samples were mechanically removed prior to analysis, then about 1 g of material per sample was manualy powdered in an agate mortar.
+
+The data were collected at room temperature with a D8 Advance (Bruker) diffractometer in Bragg-Brentano configuration working at 1.6 kW (40 kV, 40 mA) and equipped with a copper anode source ($k_{\alpha1}$ = 1.5406 ; the $k_\beta$ ray being removed by a Ni-filter in the diffracted beam).
+An 8 mm anti-scattering slit was mounted in front of the LynxEye© CCD detector.
+The explored area covered the 3-70° ($2\theta$) range, with an angle step of 0.02° and a time step of 2 seconds (the sample being rotated during the analysis).
+The stability of the instrument was checked between the different series of measurements by analyzing a standard (corundum crystal, NIST 1976).
+
+All diffractograms were processed in the same way:
+
+1. $k_{\alpha2}$ lines were stripped using penalized likelihood [@derooi2014];
+1. All diffractograms were checked for possible sample displacements, due to inaccurate powder sample mounting, then shifted accordingly (using quartz as an internal standard);
+1. All diffractograms were realigned to a common angular scale, using linear interpolation between data points;
+1. Baselines were estimated using iterative mean suppression [4S Peak Filling, @liland2015], then removed (negative values due to baseline subtraction were replaced by zeros);
+1. Data points below 4° ($2\theta$) and above 54° ($2\theta$) were removed.
+
+## Correspondence analysis
+
+We will only briefly recall the main aspects of correspondence analysis, as a detailed review of its mathematical properties is beyond the scope of this article [for an in-depth discussion, see @greenacre2007; @lebart2006].
+
+Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows to describe the statistical relationships that can exist between individuals and variables.
+Unlike PCA, CA analyses the differences between relative values and it considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space).
+CA studies the inertia of the scatterplot (i.e.
+the weighted sum of the $\chi2$ distances of each point to the centroid) to create orthogonal components, so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the component, and so on until the last dimension.
+Instead of maximizing variance, CA maximizes the correspondence between individuals and variables.
+
+## Computational environment
+
+```{r}
+#| label: bib-packages
+# Créer automatiquement une bibliographie des packages R utilisés
+dep <- here::here("DESCRIPTION") |> readLines()
+dep <- dep[seq(from = grep(pattern = "Imports", x = dep) + 1, 
+               to = max(grep(pattern = "^[[:space:]]", x = dep)))]
+dep <- gsub("[[:space:][:punct:][:digit:]]*", "", dep)
+dep <- c(dep, "base") # Add R base
+
+cite <- lapply(
+  X = dep,
+  FUN = function(x) {
+    bib <- try(citation(x), silent = FALSE)
+    if (inherits(bib, "try-error")) return(NULL)
+    entry <- utils::toBibtex(bib)
+    entry[1] <- sub("\\{,$", sprintf("{R-%s,", x), entry[1])
+    entry
+  }
+)
+here::here("analysis/paper/packages.bib") |> 
+  cat(unlist(cite), file = _, sep = "\n")
+
+pkg <- vapply(
+  X = dep[dep != "base"], 
+  FUN = function(x) {
+    sprintf("`%s` %s [@R-%s]", x, packageVersion(x), x)
+  }, 
+  FUN.VALUE = character(1)
+)
+```
+
+All the data processing and statistical analysis were performed with `r R.version$version.string` [@R-base] and the following packages: `r paste0(pkg, collapse = ", ")`.
+The R code is openly available in @frerebeau2023a and allows to replicate all the results presented below.
 

### 2023-05-19 analysis/paper/paper.qmd
Draft results

@@
 
-This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data, which aims to provide an alternative to the EFT framework.
+This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data.
+It aims to provide an alternative to the EFT framework.
 This method is based on the principle of parsimony with two main objectives: (1) to avoid the use of external references (whether experimental or ethnographic data); (2) to build the simplest possible model by limiting the number of hypotheses involved.
@@
 XRD analysis of archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing.
-Quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content.
+However, quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content.
 The estimation of quantitative mineralogical compositions, by the RIR method [@hubbard1988] or the Rietvelt refinement [@rietveld1969], is a very difficult, high skilled exercise, assuming a thorough knowledge of the material (and expensive databases).
@@
 In this study, we show that it is possible to successfully use correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials.
-The properties of correspondence analysis and its links with matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers.
+The properties of correspondence analysis and its application for matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers.
 Thus, calibration data sets can be produced on the only basis of archaeological evidence to assess ceramic firing practices across a cultural area, avoiding the use of laboratory references.
@@
 
+Finally, all the diffractograms were combined into a matrix, where each row corresponds to a sample and each column to a $2\theta$ position, so that each cell contains the corresponding X-ray photon count.
+
 ## Correspondence analysis
@@
 Unlike PCA, CA analyses the differences between relative values and it considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space).
-CA studies the inertia of the scatterplot (i.e.
+CA studies the inertia (i.e.
 the weighted sum of the $\chi2$ distances of each point to the centroid) to create orthogonal components, so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the component, and so on until the last dimension.
-Instead of maximizing variance, CA maximizes the correspondence between individuals and variables.
+CA maximizes the correspondence between individuals and variables, instead of maximizing the amount of variance explained by its reduced-space.
+
+CA is an effective method for the seriation of archaeological assemblages (REF).
+The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation.
+The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time.
+CA is not limited to chronological modeling, but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e.
+any aspect which is expected to be related to the observed composition (be it temporal or environmental).
 
@@
 pkg <- vapply(
-  X = dep[dep != "base"], 
+  X = dep[dep != "base" & dep != "here"], 
   FUN = function(x) {
@@
 # RESULTS
+
+```{r}
+#| label: xrd-data
+## Read XRD data 
+## (see analysis/supplementary-materials/01-data.R)
+xrd_counts <- here::here("analysis/data/derived_data/xrd_clean.csv") |> 
+  read.table(sep = ",", dec = ".", header = TRUE)
+
+## Remove the first column (2 theta positions)
+theta_scale  <- xrd_counts[, 1]
+xrd_counts <- xrd_counts[, -1]
+
+## Transpose (one sample per line)
+xrd_counts <- xrd_counts |> t() |> as.data.frame()
+colnames(xrd_counts) <- paste0("T", seq_len(ncol(xrd_counts)))
+
+## Get index of XRD columns
+xrd_col <- which(grepl(pattern = "T", colnames(xrd_counts)))
+
+## Read mineral data
+minerals <- here::here("analysis/data/derived_data/minerals.csv") |>
+  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8") |> 
+  subset(symbole %in% c("An", "Cal", "Gh", "Ilt", "Qz", "Di"))
+# minerals <- minerals[!duplicated(minerals$symbole), ]
+```
+
+```{r}
+#| label: ca-moreno
+## Mas de Moreno: ceramics + unfired sherds
+sup_row_id <- match(corpus$sample[corpus$site != "Mas de Moreno"], rownames(xrd_counts))
+xrd_moreno_ca <- dimensio::ca(object = xrd_counts[, xrd_col], sup_row = sup_row_id)
+
+## Get CA results
+xrd_moreno_eig <- dimensio::get_eigenvalues(xrd_moreno_ca)
+xrd_moreno_row_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 1)
+xrd_moreno_col_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 2)
+xrd_moreno_col_contrib <- dimensio::get_contributions(xrd_moreno_ca, margin = 2)
+```
+
+The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca).
+
+```{r}
+#| label: fig-plot-ca
+#| fig-cap: Scree plot of inertia (top left; only the first ten components are displayed).
+#| fig-width: 10
+#| fig-height: 10
+## Plot eigenvalues
+par(mar = c(4, 4, 1, 1), mfrow = c(2, 2))
+barplot(
+  height = xrd_moreno_eig[1:10, 1],
+  names.arg = rownames(xrd_moreno_eig)[1:10],
+  ylab = "Inertia (%)", las = 1
+)
+
+## Plot cumulative inertia
+plot(
+  x = 1:10,
+  y = xrd_moreno_eig[1:10, 3],
+  type = "b",
+  pch = 16,
+  col = "red",
+  xaxt = "n",
+  ylab = "Cumulative inertia (%)", las = 1
+)
+axis(side = 1, at = 1:10, labels = rownames(xrd_moreno_eig)[1:10])
+
+## Active rows (F1 vs F2)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = "black"
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+
+## Columns (F1 vs F2)
+plot(
+  x = xrd_moreno_col_coord[, 1],
+  y = xrd_moreno_col_coord[, 2],
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = adjustcolor("black", alpha = 0.2)
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+```
+
+```{r}
+#| label: fig-plot-contrib
+#| fig-cap: Columns contribution to the construction of the first two components. The thirty variables contributing the most to each component are marked by red lines. 
+#| fig-width: 10
+#| fig-height: 5
+par(mar = c(0, 4, 1, 1), mfrow = c(2, 1), oma = c(3, 0, 0, 0))
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 1],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F1",
+  xaxt = "n"
+)
+abline( #21.94, 27.76-27.9, 29.42
+  v = theta_scale[order(xrd_moreno_col_contrib[, 1], decreasing = TRUE)[1:30]],
+  col = adjustcolor(col = "red", alpha = 0.1)
+)
+# alkahest::peaks_find(
+#   x = theta_scale, 
+#   y = xrd_moreno_col_contrib[, 1], 
+#   SNR = 2, 
+#   m = 5
+# )
+
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 2],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F2"
+)
+abline( #12.32, 26.66, 31.36
+  v = theta_scale[order(xrd_moreno_col_contrib[, 2], decreasing = TRUE)[1:30]],
+  col = adjustcolor(col = "red", alpha = 0.1)
+)
+# alkahest::peaks_find(
+#   x = theta_scale,
+#   y = xrd_moreno_col_contrib[, 2],
+#   SNR = 2,
+#   m = 5
+# )
+mtext(expression(2*theta), side = 1, line = 2, outer = TRUE)
+```
+
+```{r}
+#| label: fig-plot-ca-loi
+#| fig-cap: Row (left) and columns (right) coordinates along the first two CA components. For the sake of readability, only the three main peaks of a selection of minerals of interest are displayed.
+#| fig-width: 10
+#| fig-height: 10
+layout(matrix(data = c(1, 2, 3, 3), ncol = 2))
+par(mar = c(4, 4, 1, 1))
+
+id_moreno <- match(rownames(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]), corpus$sample)
+
+## F1 vs F2: active only, by material
+col_material <- khroma::color("high contrast")(2)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = col_material[as.factor(corpus$material[id_moreno])]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+legend("bottomright", legend = levels(as.factor(corpus$material[id_moreno])),
+       pch = 16, col = col_material)
+
+## F1 vs F2: active only, by loi
+loi_moreno <- corpus$LOI[id_moreno]
+loi_NA <- is.na(loi_moreno)
+loi_moreno[loi_NA] <- min(loi_moreno, na.rm = TRUE)
+col_loi <- khroma::ramp(loi_moreno, palette = "YlOrBr")(5)
+col_loi[loi_NA] <- "darkgrey"
+attr(col_loi,"breaks")[loi_NA] <- NA
+loi_moreno[loi_NA] <- NA
+
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 21,
+  las = 1, asp = 1,
+  col = "black",
+  bg = col_loi
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+
+## F1 vs LOI: active only
+data_fit <- data.frame(
+  loi = loi_moreno,
+  F1 = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1]
+)
+loi_fit <- lm(loi ~ poly(F1, 2), data = data_fit)
+seq_F1 <- seq(from = min(data_fit$F1), to = max(data_fit$F1), by = 0.1)
+loi_pred <- predict(
+  object = loi_fit, 
+  newdata = data.frame(F1 = seq_F1),
+  interval = "confidence", 
+  level = 0.95
+)
+
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = loi_moreno,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = "LOI (%)",
+  type = "p",
+  pch = 21,
+  las = 1,
+  col = "black",
+  bg = col_loi
+)
+polygon(
+  x = c(seq_F1, rev(seq_F1)),
+  y = c(loi_pred[, "lwr"], rev(loi_pred[, "upr"])),
+  col = adjustcolor("black", alpha = 0.2),
+  border = NA,
+  lwd = 2
+)
+lines(
+  x = seq_F1,
+  y = loi_pred[, "fit"], 
+  col = "red",
+  lwd = 2
+)
+
+legend_image <- as.raster(sort(col_loi[col_loi != "darkgrey"]))
+rasterImage(legend_image, 0.80, 15, 0.90, 20)
+rect(0.80, 15, 0.90, 20)
+text(
+  x = 0.9, 
+  y = c(15, 20), 
+  labels = paste0(round(range(attr(col_loi, "breaks"), na.rm = TRUE)), "%"), 
+  pos = 4
+)
+```
+
+```{r}
+#| label: fig-plot-xrd
+#| fig-width: 10
+#| fig-height: 5
+par(mfrow = c(3, 1), mar = c(1, 0, 1, 1), oma = c(4, 5, 0, 0))
+
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX14922", ] / 1000,
+  type = "l",
+  xaxt = "n",
+  ylab = "",
+  las = 1
+)
+text(x = min(theta_scale) + 1, y = 55, labels = "BDX14922", font = 2)
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX15036", ] / 1000,
+  type = "l",
+  xaxt = "n",
+  ylab = "",
+  las = 1
+)
+text(x = min(theta_scale) + 1, y = 35, labels = "BDX15036", font = 2)
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX22203", ] / 1000,
+  type = "l",
+  ylab = "",
+  las = 1
+)
+text(x = min(theta_scale) + 1, y = 15.5, labels = "BDX22203", font = 2)
+
+mtext(expression(2*theta), side = 1, line = 3, cex = 1, outer = TRUE)
+mtext(expression("Count ("*10^3*")"), side = 2, line = 3, cex = 1, outer = TRUE)
+```
+
+```{r}
+#| label: ca-ceramics
+## Ceramics
+names_ok <- xrd_moreno_row_coord |> 
+  subset(F2 < 0) |> 
+  rownames()
+
+# id_OK <- match(names_ok, rownames(xrd_counts))
+id_ceram <- match(corpus$sample[corpus$material != "clay"], rownames(xrd_counts))
+xrd_ceram <- xrd_counts[id_ceram, ]
+sup_row_id <- match(corpus$sample[corpus$site != "Mas de Moreno"], rownames(xrd_ceram))
+xrd_ceram_ca <- dimensio::ca(object = xrd_ceram, sup_row = sup_row_id)
+
+## Get CA results
+xrd_ceram_eig <- dimensio::get_eigenvalues(xrd_ceram_ca)
+xrd_ceram_row_coord <- dimensio::get_coordinates(xrd_ceram_ca, margin = 1)
+xrd_ceram_col_coord <- dimensio::get_coordinates(xrd_ceram_ca, margin = 2)
+xrd_ceram_col_contrib <- dimensio::get_contributions(xrd_ceram_ca, margin = 2)
+```
+
+```{r}
+#| label: fig-plot-minerals
+#| fig-cap: XXX.
+#| fig-width: 7
+#| fig-height: 7
+theta_minerals <- vapply(
+  X = minerals$theta,
+  FUN = function(x, theta) { which.min(abs(theta - x)) },
+  FUN.VALUE = numeric(1),
+  theta = theta_scale
+)
+xrd_minerals_coord <- xrd_ceram_col_coord[theta_minerals, ]
+xlim <- range(range(xrd_ceram_row_coord[, 1]), range(xrd_minerals_coord[, 1]))
+ylim <- range(range(xrd_ceram_row_coord[, 2]), range(xrd_minerals_coord[, 2]))
+
+## Columns (F1 vs F2)
+par(mar = c(4, 4, 1, 1))
+col_minerals <- khroma::color("bright")(6)
+plot(
+  x = xrd_minerals_coord[, 1],
+  y = xrd_minerals_coord[, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = col_minerals[as.factor(minerals$symbole)]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(
+  x = xrd_minerals_coord[, 1],
+  y = xrd_minerals_coord[, 2],
+  labels = mapply(
+    FUN = function(s, h, k, l) {
+      if (h < 0) as.expression(bquote(.(s)~"("*bar(.(abs(h)))*.(k)*.(l)*")"))
+      else as.expression(bquote(.(s)~"("*.(h)*.(k)*.(l)*")"))
+    }, 
+    s = minerals$symbole,
+    h = minerals$h, 
+    k = minerals$k,
+    l = minerals$l
+  ),
+  col = col_minerals[as.factor(minerals$symbole)],
+  pos = c(3, 2, 1, 1, 2, 4, 2, 2, 2, 4, 3, 3, 1, 3, 1, 4)
+)
+```
+
+```{r}
+#| label: fig-plot-ca-ceram
+#| fig-cap: XXX.
+#| fig-width: 10
+#| fig-height: 10
+layout(matrix(data = c(1, 2, 3, 3), ncol = 2))
+par(mar = c(4, 4, 1, 1))
+
+id_moreno <- match(rownames(xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, ]), corpus$sample)
+id_settlements <- match(rownames(xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, ]), corpus$sample)
+
+## F1 vs F2: active only, by site
+col_site <- khroma::color("bright")(2)
+plot(
+  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
+  y = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
+  type = "p",
+  pch = 21,
+  las = 1, asp = 1,
+  bg = "lightgrey",
+  col = "black"
+)
+points(
+  x = xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 1],
+  y = xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 2],
+  pch = 24,
+  bg = col_site[as.factor(corpus$site[corpus$site != "Mas de Moreno"])]
+)
+# car::dataEllipse(
+#   x = as.matrix(xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 1:2]),
+#   groups = as.factor(corpus$site[id_settlements]),
+#   level = 0.95,
+#   col = col_site,
+#   plot.points = FALSE,
+#   center.pch = NULL,
+#   add = TRUE
+# )
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+legend("bottomright", legend = levels(as.factor(corpus$site[id_settlements])),
+       pch = 16, col = col_site)
+
+## F1 vs F2: active only, by loi
+loi_moreno <- corpus$LOI[id_moreno]
+loi_NA <- is.na(loi_moreno)
+loi_moreno[loi_NA] <- min(loi_moreno, na.rm = TRUE)
+col_loi <- khroma::ramp(loi_moreno, palette = "YlOrBr")(5)
+col_loi[loi_NA] <- "darkgrey"
+attr(col_loi,"breaks")[loi_NA] <- NA
+loi_moreno[loi_NA] <- NA
+
+plot(
+  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
+  y = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
+  type = "p",
+  pch = 21,
+  las = 1, asp = 1,
+  col = "black",
+  bg = col_loi
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+
+## F1 vs LOI: active only
+data_fit <- data.frame(
+  loi = loi_moreno,
+  F1 = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1]
+)
+loi_fit <- lm(loi ~ poly(F1, 2), data = data_fit)
+seq_F1 <- seq(from = min(data_fit$F1), to = max(data_fit$F1), by = 0.1)
+loi_pred <- predict(
+  object = loi_fit, 
+  newdata = data.frame(F1 = seq_F1),
+  interval = "confidence", 
+  level = 0.95
+)
+
+plot(
+  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
+  y = loi_moreno,
+  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
+  ylab = "LOI (%)",
+  type = "p",
+  pch = 21,
+  las = 1,
+  col = "black",
+  bg = col_loi
+)
+polygon(
+  x = c(seq_F1, rev(seq_F1)),
+  y = c(loi_pred[, "lwr"], rev(loi_pred[, "upr"])),
+  col = adjustcolor("black", alpha = 0.2),
+  border = NA,
+  lwd = 2
+)
+lines(
+  x = seq_F1,
+  y = loi_pred[, "fit"], 
+  col = "red",
+  lwd = 2
+)
+
+legend_image <- as.raster(sort(col_loi[col_loi != "darkgrey"]))
+rasterImage(legend_image, 0.80, 10, 0.90, 15)
+rect(0.80, 10, 0.90, 15)
+text(
+  x = 0.9, 
+  y = c(10, 15), 
+  labels = paste0(round(range(attr(col_loi, "breaks"), na.rm = TRUE)), "%"), 
+  pos = 4
+)
+```
 

### 2023-05-23 analysis/paper/paper.qmd
Write results

--- 2023-05-18 analysis/paper/paper.qmd
@@
     documentclass: scrartcl
+    include-in-header: 
+      - text: |
+          \usepackage{booktabs}
     classoption: 
@@
 
-Firing is a "strategic step" [@lemonnier1983a] of the ceramic production process, especially as it involves an economic risk: that of losing all or part of the load to be fired and as much human labour time.
-The study of the firing of archaeological ceramics is thus a key factor in the interpretation of ancient techniques and the related social practices.
-To this end, important methodological research has been carried out since the pioneering work of @shepard1936.
-This led to the wide use of the so-called *Equivalent Firing Temperatures* (EFT) as a way to describe the firing of archaeological ceramics [@tite1969; @tite1969a].
-The EFT is a point estimator tied to standard experimental conditions: "[EFT] is that temperature maintained for one hour which would produce the observed mineralogy or microstructure" [@tite1995].
+Firing is a "strategic step" [@lemonnier1983a] of the ceramic production process, especially as it involves an economic risk: the risk of losing all or part of the load to be fired and the associated human labor time.
+The study of the firing of archaeological ceramics is thus a key factor in interpreting ancient techniques and related social practices.
+In this regard, significant methodological research has been carried out since the pioneering work of @shepard1936, leading to the widespread use of the so-called *Equivalent Firing Temperatures* (EFT) as a way to describe the firing of archaeological ceramics [@tite1969; @tite1969a].
+The EFT is a point estimator tied to standard experimental conditions, defined as "the temperature maintained for one hour which would produce the observed mineralogy or microstructure" [@tite1995].
 
-The EFT is certainly practical, but it raises several methodological and conceptual problems.
+The EFT is undoubtedly practical, but it raises several methodological and conceptual problems.
 A full review is beyond the scope of this study, but two points can be made.
 On the one hand, EFT creates unnecessary ambiguity about the real firing temperature (assuming that the firing temperatures have archaeological significance).
-On the other hand, @gosselain1992 and @livingstonesmith2001 have demonstrated the limits of this framework, which may lead to anachronistic reasoning [@frerebeau2018].
+On the other hand, @gosselain1992 and @livingstonesmith2001 have demonstrated the limits of this framework, which may lead to anachronistic reasoning [see @frerebeau2018, for a discussion].
 
-This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data, which aims to provide an alternative to the EFT framework.
-This method is based on the principle of parsimony with two main objectives: (1) to avoid the use of external references (whether experimental or ethnographic data); (2) to build the simplest possible model by limiting the number of hypotheses involved.
+Therefore, this study proposes a new method based on the multivariate analysis of powder X-ray diffraction (XRD) data, aiming to provide an alternative to the EFT framework.
+This method follows the principle of parsimony and has two main objectives: (1) to avoid the use of external references, whether experimental or ethnographic data; (2) to construct the simplest possible model by limiting the number of hypotheses involved.
 
-XRD analysis of archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing.
-Quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content.
-The estimation of quantitative mineralogical compositions, by the RIR method [@hubbard1988] or the Rietvelt refinement [@rietveld1969], is a very difficult, high skilled exercise, assuming a thorough knowledge of the material (and expensive databases).
+XRD analysis of archaeological ceramics allows for the identification of the main minerals, with special attention paid to the presence or absence of mineralogical phases formed during firing.
+However, quantitative analysis of XRD data faces significant difficulties due to the multiphase nature, poor crystallization, and amorphous content of archaeological ceramics.
+Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material and access to expensive databases.
 
-In this study, we show that it is possible to successfully use correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials.
-The properties of correspondence analysis and its links with matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers.
-Thus, calibration data sets can be produced on the only basis of archaeological evidence to assess ceramic firing practices across a cultural area, avoiding the use of laboratory references.
+In this study, we demonstrate the successful use of correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials.
+The properties of correspondence analysis and its application for matrix seriation enable the characterization of the firing of a ceramic assemblage and identification of outliers.
+Calibration data sets can be produced solely based on archaeological evidence to assess ceramic firing practices, eliminating the need for laboratory references.
 
-As an illustration of this method, we examine XRD data of Iron Age ceramics from an Iberian workshop.
-The Mas de Moreno workshop is located in the province of Teruel (Aragon, Spain), in the municipality of Foz-Calanda.
-The activity of the workshop spans between 225-200 BC and 40-30 AD.
-A first archaeological synthesis of the workshop has highlighted three successive phases of activity, characterised by successive changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012].
-Several kilns have been excavated, which were discontinued and destroyed around 50-40 BC [@gorgues2012].
-The use of an amphora kiln, of much larger dimensions than the previous kilns, as well as the production of Roman amphora and the presence of Latin epigraphy, illustrate a major reorganisation of the workshop in the middle of the 1^st^ century.
-The workshop's activity ceased during the following decade, mostly as a result of the territorial reorganisation under Roman influence and the distance from the trade routes [@gorgues2012].
-This workshop thus constitutes a context well suited to investigate the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.
+To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop, located in the province of Teruel (Aragon, Spain), in the municipality of Foz-Calanda.
+The workshop was active between 225-200 BC and 40-30 AD.
+Previous archaeological studies of the workshop have highlighted three successive phases of activity, characterized by changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012].
+Several kilns have been excavated, with discontinuation and destruction occurring around 50-40 BC [@gorgues2012].
+The use of a larger amphora kiln, the production of Roman amphorae, and the presence of Latin epigraphy indicate a major reorganization of the workshop in the middle of the 1st century.
+The workshop ceased its activity in the following decade due to territorial reorganization under Roman influence and its distance from trade routes [@gorgues2012].
+This workshop provides a context well-suited for investigating the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.
 
@@
 #| label: tbl-corpus
-#| tbl-cap: "Ceramic and clay samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. LOI: loss on ignition [see @frerebeau2020a]."
+#| tbl-cap: "Ceramic and clay samples analyzed in this study.
+Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC.
+SU: stratigraphic unit.
+LOI: loss on ignition [see @frerebeau2020a]."
+corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration", "part"))] |> 
   knitr::kable(
-  x = corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration"))],
-  col.names = c("Sample", "Material", "Site", "Stratigraphic Unit", "Phase", "Type", "Part", "LOI (%)")
+    col.names = c("Sample", "Material", "Site", "SU", "Phase", "Type", "LOI (%)"),
+    booktabs = TRUE,
+    longtable = TRUE
   )
@@
 ## Powder X-ray diffraction
+
+```{r}
+#| label: xrd-data
+## Read XRD data 
+## (see analysis/supplementary-materials/01-data.R)
+xrd_counts <- here::here("analysis/data/derived_data/xrd_clean.csv") |> 
+  read.table(sep = ",", dec = ".", header = TRUE)
+
+## Remove the first column (2 theta positions)
+theta_scale  <- xrd_counts[, 1]
+xrd_counts <- xrd_counts[, -1]
+
+## Transpose (one sample per line)
+xrd_counts <- xrd_counts |> t() |> as.data.frame()
+colnames(xrd_counts) <- paste0("T", seq_len(ncol(xrd_counts)))
+
+## Get index of XRD columns
+xrd_col <- which(grepl(pattern = "T", colnames(xrd_counts)))
+
+## Read mineral data
+minerals <- here::here("analysis/data/derived_data/minerals.csv") |>
+  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8") |> 
+  subset(symbole %in% c("An", "Cal", "Gh", "Ilt", "Qz", "Di"))
+# minerals <- minerals[!duplicated(minerals$symbole), ]
+```
+
 The mineralogical composition of the samples was determined by X-ray powder diffraction.
-The outer surfaces of all the samples were mechanically removed prior to analysis, then about 1 g of material per sample was manualy powdered in an agate mortar.
+The outer surfaces of all the samples were mechanically removed prior to analysis.
+Then, approximately 1 g of material per sample was manually powdered in an agate mortar.
 
-The data were collected at room temperature with a D8 Advance (Bruker) diffractometer in Bragg-Brentano configuration working at 1.6 kW (40 kV, 40 mA) and equipped with a copper anode source ($k_{\alpha1}$ = 1.5406 ; the $k_\beta$ ray being removed by a Ni-filter in the diffracted beam).
+The data were collected at room temperature using a D8 Advance (Bruker) diffractometer in Bragg-Brentano configuration operating at 1.6 kW (40 kV, 40 mA) and equipped with a copper anode source ($k_{\alpha1}$ = 1.5406 Å; the $k_\beta$ ray was removed by a Ni-filter in the diffracted beam).
 An 8 mm anti-scattering slit was mounted in front of the LynxEye© CCD detector.
-The explored area covered the 3-70° ($2\theta$) range, with an angle step of 0.02° and a time step of 2 seconds (the sample being rotated during the analysis).
-The stability of the instrument was checked between the different series of measurements by analyzing a standard (corundum crystal, NIST 1976).
+The explored area covered the 3-70° ($2\theta$) range, with an angle step of 0.02° and a time step of 2 seconds (the sample was rotated during the analysis).
+The stability of the instrument was checked between different series of measurements by analyzing a standard (corundum crystal, NIST 1976).
 
@@
 
-1. $k_{\alpha2}$ lines were stripped using penalized likelihood [@derooi2014];
-1. All diffractograms were checked for possible sample displacements, due to inaccurate powder sample mounting, then shifted accordingly (using quartz as an internal standard);
-1. All diffractograms were realigned to a common angular scale, using linear interpolation between data points;
-1. Baselines were estimated using iterative mean suppression [4S Peak Filling, @liland2015], then removed (negative values due to baseline subtraction were replaced by zeros);
-1. Data points below 4° ($2\theta$) and above 54° ($2\theta$) were removed.
+* $k_{\alpha2}$ lines were stripped using penalized likelihood [@derooi2014].
+* All diffractograms were checked for possible sample displacements due to inaccurate powder sample mounting and then shifted accordingly (using quartz as an internal standard).
+* All diffractograms were realigned to a common angular scale using linear interpolation between data points.
+* Baselines were estimated using iterative mean suppression [4S Peak Filling, @liland2015] and then removed (negative values due to baseline subtraction were replaced by zeros).
+* Data points below 4° ($2\theta$) and above 54° ($2\theta$) were removed.
+
+Finally, all the diffractograms were combined into a `r paste0(dim(xrd_counts[, xrd_col]), collapse = " x ")` matrix where each row corresponds to a sample and each column corresponds to a $2\theta$ position.
+Each cell contains the corresponding X-ray photon count.
 
@@
 
-Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows to describe the statistical relationships that can exist between individuals and variables.
-Unlike PCA, CA analyses the differences between relative values and it considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space).
-CA studies the inertia of the scatterplot (i.e.
-the weighted sum of the $\chi2$ distances of each point to the centroid) to create orthogonal components, so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the component, and so on until the last dimension.
-Instead of maximizing variance, CA maximizes the correspondence between individuals and variables.
+Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows describing the statistical relationships that can exist between individuals and variables.
+However, the concept of similarity between rows and columns is different.
+In CA, two rows (resp.
+columns) are close to each other if they associate with columns (resp.
+rows) in the same way.
+Unlike PCA, CA analyzes the differences between relative values and considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space).
+CA studies the inertia (i.e., the weighted sum of the χ2 distances of each point to the centroid) to create orthogonal components so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the second component, and so on until the last dimension.
+CA maximizes the correspondence between individuals and variables instead of maximizing the amount of variance explained by its reduced space.
+
+CA is an effective method for the seriation of archaeological assemblages (REF).
+The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation.
+The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time.
+CA is not limited to chronological modeling but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e., any aspect that is expected to be related to the observed composition (be it temporal or environmental).
 
@@
 # RESULTS
+
+```{r}
+#| label: ca-moreno
+## Mas de Moreno: ceramics + unfired sherds
+sup_row_id <- match(corpus$sample[corpus$site != "Mas de Moreno"], rownames(xrd_counts))
+xrd_moreno_ca <- dimensio::ca(object = xrd_counts[, xrd_col], sup_row = sup_row_id)
+
+## Get CA results
+xrd_moreno_eig <- dimensio::get_eigenvalues(xrd_moreno_ca)
+xrd_moreno_row_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 1)
+xrd_moreno_col_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 2)
+xrd_moreno_col_contrib <- dimensio::get_contributions(xrd_moreno_ca, margin = 2)
+```
+
+The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca.
+Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and shards.
+The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals.
+These additional individuals do not contribute to the calculation of the correspondence analysis itself, but they are projected onto the existing analysis space to visualize their relationships with the original dataset.
+
+The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca A and B).
+Plotting the coordinates of the individuals shows an interesting pattern (@fig-plot-ca C and D).
+The upper quadrants each consist of a small group of closely located individuals, likely exhibiting high similarity to one another.
+The remaining samples are distributed in the lower quadrants.
+
+```{r}
+#| label: fig-plot-ca
+#| fig-cap: "CA results of the XRD patterns of the Mas de Moreno samples.
+(A) Scree plot of inertia (only the first ten components are displayed).
+(B) Cumulative inertia of the first ten components.
+(C) CA rows score plot.
+(D) CA columns score plot."
+#| fig-width: 10
+#| fig-height: 10
+## Plot eigenvalues
+par(mar = c(4, 4, 1, 1), mfrow = c(2, 2))
+mid <- barplot(
+  height = xrd_moreno_eig[1:10, 1],
+  names.arg = rownames(xrd_moreno_eig)[1:10],
+  ylab = "Inertia (%)", las = 1, ylim = c(0, 0.3)
+)
+text(x = min(mid), y = 0.29, labels = "A", font = 2, cex = 1.5)
+
+## Plot cumulative inertia
+plot(
+  x = 1:10,
+  y = xrd_moreno_eig[1:10, 3],
+  type = "b",
+  pch = 16,
+  col = "red",
+  xaxt = "n",
+  xlab = "",
+  ylab = "Cumulative inertia (%)", las = 1
+)
+text(x = 1, y = 92, labels = "B", font = 2, cex = 1.5)
+axis(side = 1, at = 1:10, labels = rownames(xrd_moreno_eig)[1:10])
+
+## Active rows (F1 vs F2)
+xlim <- range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1], xrd_moreno_col_coord[, 1])
+ylim <- range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2], xrd_moreno_col_coord[, 2])
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = "black"
+)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "C", font = 2, cex = 1.5)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+
+## Columns (F1 vs F2)
+plot(
+  x = xrd_moreno_col_coord[, 1],
+  y = xrd_moreno_col_coord[, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = adjustcolor("black", alpha = 0.2)
+)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+```
+
+@fig-plot-contrib shows the contribution of the different columns to the construction of the first two CA-axes.
+It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix.
+Consequently, the columns that contribute the most to the construction of the first axis (with a contribution greater than 0.5) are associated with the following diffraction peaks: 21.94° (plagioclases), 26.64° (quartz), 27.76° (plagioclases), 27.90° (plagioclases), and 29.42° (calcite).
+Similarly, for the second axis, the columns that contribute the most are those corresponding to the peaks at 12.32° (phyllosilicate), 26.66° (quartz), and 31.36° (melilites).
+
+```{r}
+#| label: fig-plot-contrib
+#| fig-cap: Columns contribution to the construction of the first two components. The twenty variables contributing the most to each component are highlighted in red. 
+#| fig-width: 10
+#| fig-height: 5
+par(mar = c(0, 4, 1, 1), mfrow = c(2, 1), oma = c(3, 0, 0, 0))
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 1],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F1",
+  xaxt = "n"
+)
+abline( #21.94, 27.76-27.9, 29.42
+  v = theta_scale[order(xrd_moreno_col_contrib[, 1], decreasing = TRUE)[1:20]],
+  col = adjustcolor(col = "red", alpha = 0.1)
+)
+# pks_contrib1 <- alkahest::peaks_find(
+#   x = theta_scale,
+#   y = xrd_moreno_col_contrib[, 1],
+#   SNR = 2,
+#   m = 5
+# )
+# dput(pks_contrib1$x[pks_contrib1$y > 0.5])
+# points(x = pks_contrib1$x[pks_contrib1$y > 0.5],
+#        y = pks_contrib1$y[pks_contrib1$y > 0.5])
+
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 2],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F2"
+)
+abline( #12.32, 26.66, 31.36
+  v = theta_scale[order(xrd_moreno_col_contrib[, 2], decreasing = TRUE)[1:20]],
+  col = adjustcolor(col = "red", alpha = 0.1)
+)
+mtext(expression(2*theta), side = 1, line = 2, outer = TRUE)
+# pks_contrib2 <- alkahest::peaks_find(
+#   x = theta_scale,
+#   y = xrd_moreno_col_contrib[, 2],
+#   SNR = 2,
+#   m = 5
+# )
+# dput(pks_contrib2$x[pks_contrib2$y > 0.5])
+# points(x = pks_contrib2$x[pks_contrib2$y > 0.5],
+#        y = pks_contrib2$y[pks_contrib2$y > 0.5])
+```
+
+The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman).
+This effect is characterized by a strong nonlinear relationship between the second factor and the first factor.
+The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014].
+Interestingly, the samples located in the upper right quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A and B, @fig-plot-xrd).
+These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1.
+Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.
+
+```{r}
+#| label: fig-plot-guttman
+#| fig-cap: "Guttman effect observed on the first plane of correspondence analysis.
+Row coordinates are highlighted based on material (A), chronological phase (C), and supplementary individuals (C).
+In the top right (B), the column coordinates are shown, displaying only the main diffraction lines of a selected set of minerals of interest for improved readability.
+Ilt: illite; Cal: calcite; Qz: quartz; Gh: gehlenite; Di: diopside; An: anorthite."
+#| fig-width: 10
+#| fig-height: 10
+layout(matrix(data = c(1, 3, 2, 4), ncol = 2))
+par(mar = c(4, 4, 1, 1))
+
+theta_minerals <- vapply(
+  X = minerals$theta,
+  FUN = function(x, theta) { which.min(abs(theta - x)) },
+  FUN.VALUE = numeric(1),
+  theta = theta_scale
+)
+xrd_minerals_coord <- xrd_moreno_col_coord[theta_minerals, ]
+xlim <- range(range(xrd_moreno_row_coord[, 1]), range(xrd_minerals_coord[, 1]))
+ylim <- range(range(xrd_moreno_row_coord[, 2]), range(xrd_minerals_coord[, 2]))
+
+id_moreno <- match(rownames(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]), corpus$sample)
+
+## F1 vs F2: active only, by material
+col_material <- khroma::color("high contrast")(2)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = c(16, 17)[as.factor(corpus$material[id_moreno])],
+  las = 1, asp = 1,
+  col = col_material[as.factor(corpus$material[id_moreno])]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "A", font = 2, cex = 1.5)
+legend("bottomright", legend = levels(as.factor(corpus$material[id_moreno])),
+       pch = c(16, 17), col = col_material)
+
+## Columns (F1 vs F2)
+par(mar = c(4, 4, 1, 1))
+col_minerals <- khroma::color("bright")(6)
+plot(
+  x = xrd_minerals_coord[, 1],
+  y = xrd_minerals_coord[, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = col_minerals[as.factor(minerals$symbole)]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(
+  x = xrd_minerals_coord[, 1],
+  y = xrd_minerals_coord[, 2],
+  labels = mapply(
+    FUN = function(s, h, k, l) {
+      if (h < 0) as.expression(bquote(.(s)~"("*bar(.(abs(h)))*.(k)*.(l)*")"))
+      else as.expression(bquote(.(s)~"("*.(h)*.(k)*.(l)*")"))
+    }, 
+    s = minerals$symbole,
+    h = minerals$h, 
+    k = minerals$k,
+    l = minerals$l
+  ),
+  col = col_minerals[as.factor(minerals$symbole)],
+  pos = c(2,3, 2, 1, 4, 4, 4, 2, 2, 2, 4, 3, 3, 4, 3, 1, 4)
+)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "B", font = 2, cex = 1.5)
+
+## F1 vs F2: active only, chronology
+col_phase <- khroma::color("bright")(3)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = c(21, 22, 23)[as.factor(corpus$phase[id_moreno])],
+  las = 1, asp = 1,
+  col = "black",
+  bg = col_phase[as.factor(corpus$phase[id_moreno])]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "C", font = 2, cex = 1.5)
+legend("bottomright", 
+       legend = paste0("Phase ", levels(as.factor(corpus$phase[id_moreno]))),
+       pch = c(21, 22, 23), col = "black", pt.bg = col_phase)
+
+## F1 vs F2: active and supplementary rows
+col_site <- khroma::color("vibrant")(2)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 16,
+  las = 1, asp = 1,
+  col = "lightgrey"
+)
+points(
+  x = xrd_moreno_row_coord[xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[xrd_moreno_row_coord$.sup, 2],
+  pch = c(24, 25)[as.factor(corpus$site[corpus$site != "Mas de Moreno"])],
+  bg = col_site[as.factor(corpus$site[corpus$site != "Mas de Moreno"])]
+)
+car::dataEllipse(
+  x = as.matrix(xrd_moreno_row_coord[xrd_moreno_row_coord$.sup, 1:2]),
+  groups = as.factor(corpus$site[corpus$site != "Mas de Moreno"]),
+  group.labels = NULL,
+  level = 0.95,
+  col = col_site,
+  plot.points = FALSE,
+  center.pch = NULL,
+  add = TRUE
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
+legend("bottomright", legend = c(levels(as.factor(corpus$site[corpus$site != "Mas de Moreno"])), "Mas de Moreno"),
+       pch = c(24, 25, 16), pt.bg = col_site, col = c("black", "black", "lightgrey"))
+
+```
+
+```{r}
+#| label: fig-plot-xrd
+#| fig-cap: "Diffractograms of representative samples from each quadrant of the CA plot (see @fig-plot-ca).
+From top to bottom: upper left quadrant (BDX14922), lower left quadrant (BDX15451), lower right quadrant (BDX15465), upper right quadrant (BDX22203).
+Only the peaks that contribute the most to the construction of the first two CA axes are indexed (see text and @fig-plot-contrib)."
+#| fig-width: 10
+#| fig-height: 11
+par(mfrow = c(4, 1), mar = c(1, 0, 1, 1), oma = c(4, 4, 4, 0))
+
+pks <- c(c(21.94, 26.64, 27.76, 27.9, 29.42), c(12.32, 26.66, 31.36))
+## Upper left quadrant
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX14922", ] / 1000,
+  type = "l",
+  xaxt = "n",
+  yaxt = "n",
+  ylab = "",
+  las = 1
+)
+abline(v = pks, lty = 2)
+axis(side = 3, at = pks)
+text(x = min(theta_scale) + 1, y = 55, labels = "BDX14922", font = 2)
+## Bottom left quadrant
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX15451", ] / 1000,
+  type = "l",
+  xaxt = "n",
+  yaxt = "n",
+  ylab = "",
+  las = 1
+)
+abline(v = pks, lty = 2)
+text(x = min(theta_scale) + 1, y = 35, labels = "BDX15451", font = 2)
+## Bottom right quadrant
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX15465", ] / 1000,
+  type = "l",
+  xaxt = "n",
+  yaxt = "n",
+  ylab = "",
+  las = 1
+)
+abline(v = pks, lty = 2)
+text(x = min(theta_scale) + 1, y = 35, labels = "BDX15465", font = 2)
+## Top right quadrant
+plot(
+  x = theta_scale,
+  y = xrd_counts[rownames(xrd_counts) == "BDX22203", ] / 1000,
+  type = "l",
+  yaxt = "n",
+  ylab = "",
+  las = 1
+)
+abline(v = pks, lty = 2)
+text(x = min(theta_scale) + 1, y = 15.5, labels = "BDX22203", font = 2)
+
+mtext(expression(2*theta), side = 1, line = 3, cex = 1, outer = TRUE)
+mtext("Count (a. u.)", side = 2, line = 2, cex = 1, outer = TRUE)
+```
+
+```{r}
+#| label: fig-plot-ca-loi
+#| fig-cap: "Bottom right: row coordinates along the first CA-axis according to the loss on ignition (LOI)."
+#| fig-width: 10
+#| fig-height: 5
+par(mar = c(4, 4, 1, 1), mfrow = c(1, 2), oma = c(0, 0, 0, 0))
+
+loi_moreno <- corpus$LOI[id_moreno]
+loi_NA <- is.na(loi_moreno)
+loi_moreno[loi_NA] <- min(loi_moreno, na.rm = TRUE)
+col_loi <- khroma::ramp(loi_moreno, palette = "YlOrBr")(5)
+col_loi[loi_NA] <- "darkgrey"
+attr(col_loi,"breaks")[loi_NA] <- NA
+loi_moreno[loi_NA] <- NA
+
+## F1 vs F2: LOI
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = 21,
+  las = 1, asp = 1,
+  col = "black",
+  bg = col_loi
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+
+## F1 vs LOI
+data_fit <- data.frame(
+  LOI = loi_moreno,
+  F1 = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1]
+)
+data_fit <- data_fit[stats::complete.cases(data_fit), ] # Remove NA
+loi_fit <- lm(LOI ~ poly(F1, 2), data = data_fit)
+loi_fit_res <- summary(loi_fit)
+seq_F1<- seq(from = min(data_fit$F1), to = max(data_fit$F1), by = 0.1)
+loi_pred <- predict(
+  object = loi_fit, 
+  newdata = data.frame(F1 = seq_F1),
+  interval = "confidence", 
+  level = 0.95
+)
+
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = loi_moreno,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = "LOI (%)",
+  type = "p",
+  pch = 21,
+  las = 1,
+  col = "black",
+  bg = col_loi
+)
+polygon(
+  x = c(seq_F1, rev(seq_F1)),
+  y = c(loi_pred[, "lwr"], rev(loi_pred[, "upr"])),
+  col = adjustcolor("black", alpha = 0.2),
+  border = NA,
+  lwd = 2
+)
+lines(
+  x = seq_F1,
+  y = loi_pred[, "fit"], 
+  col = "red",
+  lwd = 2
+)
+
+legend_image <- as.raster(sort(col_loi[col_loi != "darkgrey"]))
+rasterImage(legend_image, 0.80, 15, 0.90, 20)
+rect(0.80, 15, 0.90, 20)
+text(
+  x = 0.9, 
+  y = c(15, 20), 
+  labels = paste0(round(range(attr(col_loi, "breaks"), na.rm = TRUE)), "%"), 
+  pos = 4
+)
+```
+
+Additionally, @fig-plot-ca-loi illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples.
+It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis.
+To further explore this relationship, a second-order polynomial regression analysis was conducted.
+The results indicated that the overall regression was statistically significant (RSE = `r round(loi_fit_res$sigma, 2)` on `r  loi_fit_res$df[2]` degrees of freedom, R^2^ = `r round(loi_fit_res$r.squared, 2)`, adjusted R^2^ = `r round(loi_fit_res$adj.r.squared, 2)`, `r sprintf("F(%g, %g) = %g", loi_fit_res$fstatistic[2], loi_fit_res$fstatistic[3], round(loi_fit_res$fstatistic[1], 1))`, p < 0.001), indicating a good fit of the polynomial regression model to the data.
+
+The Guttman effect arises from the unimodal distribution of observed variables along a gradient, suggesting the possibility of ordering these variables along the same gradient.
+Our results strongly suggest that the gradient under consideration is related to the firing process: the first axis represents the opposition between minerals present in the raw material and those formed during firing.
+These secondary minerals gradually form during the firing process as conditions change and are likely to be consumed in subsequent chemical reactions.
+Hence, a unimodal distribution can be expected for various $2\theta$ positions along this gradient.
 

### 2023-05-23 analysis/paper/paper.qmd
Update introduction

@@
 
-Firing is a "strategic step" [@lemonnier1983a] of the ceramic production process, especially as it involves an economic risk: the risk of losing all or part of the load to be fired and the associated human labor time.
+Firing is a "strategic step" [@lemonnier1983a] of the ceramic production process: its success largely determines the final mechanical, refractory and --- to some extent --- aesthetic properties of the artifacts.
+Moreover, in case of failure this step involves the risk of losing all or part of the load to be fired and the associated human labor time.
 The study of the firing of archaeological ceramics is thus a key factor in interpreting ancient techniques and related social practices.
@@
 The EFT is undoubtedly practical, but it raises several methodological and conceptual problems.
-A full review is beyond the scope of this study, but two points can be made.
-On the one hand, EFT creates unnecessary ambiguity about the real firing temperature (assuming that the firing temperatures have archaeological significance).
-On the other hand, @gosselain1992 and @livingstonesmith2001 have demonstrated the limits of this framework, which may lead to anachronistic reasoning [see @frerebeau2018, for a discussion].
+A full review is beyond the scope of this study [see @frerebeau2018 for a detailed discussion], but two points can be made.
+On the one hand, EFT creates unnecessary ambiguity about the real firing temperature.
+Tite [-@tite1995; -@tite1999] points out that the EFT does not account for the temperature within the firing structure.
+At best, it serves as an indication of the heat transmitted during the firing process that led to the observed transformations (mineralogical or microstructural), i.e., a combination of temperature and time.
+On the other hand, @gosselain1992 and @livingstonesmith2001 have highlighted the limitations of this framework.
+EFT alone does not provide a precise understanding of the firing conditions or the specific firing techniques employed.
 
@@
 
-XRD analysis of archaeological ceramics allows for the identification of the main minerals, with special attention paid to the presence or absence of mineralogical phases formed during firing.
-However, quantitative analysis of XRD data faces significant difficulties due to the multiphase nature, poor crystallization, and amorphous content of archaeological ceramics.
+XRD is routinely used to investigate the mineralogical composition of archaeological ceramic bodies.
+Typically, XRD is employed qualitatively, focusing on identifying mineralogical phases that have formed during firing, for understanding the firing history and the transformations that occurred within the ceramics.
+Quantitative analysis of XRD data faces significant difficulties due to the multiphase nature, poor crystallization, and amorphous content of archaeological ceramics.
 Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material and access to expensive databases.
 
-In this study, we demonstrate the successful use of correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials.
-The properties of correspondence analysis and its application for matrix seriation enable the characterization of the firing of a ceramic assemblage and identification of outliers.
+Limited work has been done on the use of multivariate methods for the analysis of ray diffraction data from archaeological materials.
+Different approaches have been used such as Lineal Discriminant Analysis [@garciagimenez2006], Principal Component Analysis [@delavilla2003; @holakooei2014] or hierarchical clustering [@piovesan2013; @maritan2015], mainly for classification purposes.
+
+In this study, we show that correspondence analysis (CA) of X-ray diffractograms can be used to assess the firing of ceramic materials.
+The properties of correspondence analysis and its application for matrix seriation enable the description of the firing of a ceramic assemblage and identification of outliers.
 Calibration data sets can be produced solely based on archaeological evidence to assess ceramic firing practices, eliminating the need for laboratory references.
 
-To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop, located in the province of Teruel (Aragon, Spain), in the municipality of Foz-Calanda.
+To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop, located in the province of Teruel (Aragon, Spain).
 The workshop was active between 225-200 BC and 40-30 AD.
@@
 
-`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop were sampled to represent the entire period of the workshop's activity (@tbl-corpus).
+`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon) were sampled to represent the entire period of the workshop's activity [stratigraphic and typological attributions can be accessed in @frerebeau2015g; @frerebeau2015e].
 The ceramic material of the Mas de Moreno consists only of products known as Iberian painted/fine wares.
@@
 LOI: loss on ignition [see @frerebeau2020a]."
+#| eval: false
 corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration", "part"))] |> 
@@
 A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017].
-Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (@tbl-corpus).
+Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here.
 
-Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (@tbl-corpus).
+Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled.
 These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dates between the 8^th^ and the 4^th^ century BC).
@@
 
-Additionally, @fig-plot-ca-loi illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples.
+Additionally, @fig-plot-ca-loi illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data].
 It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis.

### 2023-05-25 analysis/paper/paper.qmd
Write discussion and conclusion

@@
 keywords: |
-  XRD; ceramic firing; correspondence analysis; matrix seriation
+  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
 highlights: |
@@
 Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
 
@@
 XRD is routinely used to investigate the mineralogical composition of archaeological ceramic bodies.
-Typically, XRD is employed qualitatively, focusing on identifying mineralogical phases that have formed during firing, for understanding the firing history and the transformations that occurred within the ceramics.
+Typically, XRD is employed qualitatively, focusing on identifying mineralogical phases that have formed during firing, for understanding the firing history and the transformations that occurred within the ceramic material.
 Quantitative analysis of XRD data faces significant difficulties due to the multiphase nature, poor crystallization, and amorphous content of archaeological ceramics.
-Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material and access to expensive databases.
+Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material (and may require access to expensive databases).
 
@@
 
-To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop, located in the province of Teruel (Aragon, Spain).
+To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop (Teruel, Aragon, Spain).
 The workshop was active between 225-200 BC and 40-30 AD.
@@
 Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled.
-These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dates between the 8^th^ and the 4^th^ century BC).
+These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC).
 Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study.
@@
 #| fig-cap: "Guttman effect observed on the first plane of correspondence analysis.
-Row coordinates are highlighted based on material (A), chronological phase (C), and supplementary individuals (C).
+Row coordinates are highlighted based on material (A), chronological phase (C), and supplementary individuals with 95% probability ellipses (D).
 In the top right (B), the column coordinates are shown, displaying only the main diffraction lines of a selected set of minerals of interest for improved readability.
@@
 
+The Guttman effect arises from the unimodal distribution of observed variables along a gradient, suggesting the possibility of ordering these variables along the same gradient.
+Our results strongly suggest that the gradient under consideration is related to the firing process: the first axis represents the opposition between minerals present in the raw material and those formed during firing.
+These secondary minerals gradually form during the firing process as conditions change and are likely to be used in subsequent chemical reactions.
+Hence, a unimodal distribution can be expected for various $2\theta$ positions along this gradient.
+
+# DISCUSSION
+## Ceramic firing
+
+The production of ceramic materials involves solid-state transformations where the mineralogical composition of the final product differs from that of the raw material [@heimann1989].
+This is particularly true for Ca-rich ceramics, where the formation of new Ca-silicates can occur.
+The energy required for these transformations is supplied in the form of heat during firing, without reaching melting.
+As a result, the transformation process tends toward thermodynamic equilibrium but never fully achieves it [@grapes2010, 191].
+One of the reasons for this deviation from equilibrium lies in the control exerted by reaction kinetics during ceramic firing [@treiman1983; @heimann1989; @heimann2019].
+The majority of archaeological ceramic materials are characterized by frozen-in phase transitions [@rocabois2001; @tsuchiyama1983; @heimann2019].
+
+Archaeological ceramic thus constitute highly heterogeneous materials, allowing for the existence of micro-equilibria confined to the reaction interfaces between grains.
+These micro-equilibria can result in the formation of thermodynamically incompatible mineral phases at the system scale [@maggetti1986; @heimann2019].
+However, the appearance of new phases is dependent on the availability of reactive components, which delays the initiation of these transformations compared to situations described by phase diagrams.
+Numerous studies have investigated these transformations [see, for instance, @peters1978; @dondi1998; @duminuco1998; @riccardi1999; @cultrone2001; @rathossi2010], we only highlight the main findings in this context.
+
+* During the firing process, the dehydroxylation of clay minerals and the thermal decomposition of calcium carbonates facilitate the availability of highly reactive oxides, enabling the formation of new phases, whether crystalline or not.
+* At the interface between quartz and phyllosilicates, the dehydroxylation of phyllosilicates leads to the expected formation of potassium feldspars and alumino-silicates (mullite).
+Similarly, the interaction between dolomite and quartz is likely to result in the appearance of diopside.
+Furthermore, the contact between calcite and quartz is expected to produce wollastonite, which appears to be an intermediate reaction promoting the formation of anorthite [@traore2000].
+* The interface between calcite, quartz, and phyllosilicates is anticipated to yield different calcium silicates (gehlenite, anorthite, wollastonite, and pyroxenes), sometimes accompanied by phases associated with the dehydroxylation of clay minerals, particularly potassium feldspars.
+The formation of gehlenite is a phenomenon that is accentuated by local enrichment in CaO.
+
+The principal axis of correspondence analysis is built upon the opposition between primary phases and secondary phases.
+The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (@fig-plot-guttman B).
+The intermediate position of gehlenite can also be explained.
+The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle [@frerebeau2020a].
+In this case, the formed gehlenite is metastable.
+It serves as an intermediate product towards the formation of anorthite and wollastonite, linked to the diffusion of calcium through the matrix [@traore2000; @traore2003a].
+As gehlenite forms a solid solution with akermanite (the melilite series), it can also promote the crystallization of diopside (primarily of the fassaite type) rather than anorthite [@dondi1998; @rathossi2010].
+
 ```{r}
-#| label: fig-plot-ca-loi
-#| fig-cap: "Bottom right: row coordinates along the first CA-axis according to the loss on ignition (LOI)."
-#| fig-width: 10
-#| fig-height: 5
-par(mar = c(4, 4, 1, 1), mfrow = c(1, 2), oma = c(0, 0, 0, 0))
+#| label: fig-plot-loi-map
+#| fig-cap: "Top: the relationship between loss on ignition (LOI) and CA coordinates is depicted.
+The red line represents the second-order polynomial fit with an associated 95% confidence interval.
+Bottom: a heat map of the diffractograms in the 5-35° range ($2\\theta$) is displayed.
+The samples are ordered based on the first axis of the correspondence analysis.
+To enhance visualization, a square root transformation was applied."
+#| fig-width: 8
+#| fig-height: 8
+layout(matrix(c(1, 3, 3, 2, 3, 3), ncol = 2))
+par(mar = c(4, 4, 1, 2))
 
@@
 )
+
+## Heat map
+xrd_moreno_row_active <- xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]
+which_moreno <- match(rownames(xrd_moreno_row_active), rownames(xrd_counts))
+xrd_counts_moreno <- xrd_counts[which_moreno, ]
+
+id_order <- order(xrd_moreno_row_active[, 1])
+xrd_reorder <- as.matrix(xrd_counts_moreno[id_order, ])
+xrd_reorder <- t(sqrt(xrd_reorder))
+# xrd_reorder <- apply(X = xrd_reorder, MARGIN = 1, FUN = function(x) alkahest::rescale_max(seq_along(x), x)$y)
+
+theta_range <- theta_scale >= 5 & theta_scale <= 35
+image(
+  x = theta_scale[theta_range],
+  y = seq_len(ncol(xrd_reorder)),
+  z = xrd_reorder[theta_range, ],
+  xlab = expression(2*theta), ylab = "",
+  yaxt = "n",
+  col = khroma::color("YlOrBr")(255)
+)
+arrows(x0 = 5, y0 = 0, x1 = 5, y1 = 70.5, length = 0.2, lwd = 2, xpd = TRUE)
+abline(
+  h = which.min(xrd_moreno_row_active[id_order, 1] < 0 & xrd_moreno_row_active[id_order, 2] > 0) - 0.5,
+  lty = 2
+)
+abline(
+  h = which.max(xrd_moreno_row_active[id_order, 1] > 0 & xrd_moreno_row_active[id_order, 2] > 0) - 0.5,
+  lty = 2
+)
+axis(side = 4, at = c(11.5, 60.5), labels = FALSE)
+mtext(text = "firing gradient", side = 2, line = 0.5)
+mtext(text = "unfired", side = 4, line = 0.5, at = 5.5)
+mtext(text = "production conditions", line = 0.5, side = 4, at = 35)
+mtext(text = "overfired", line = 0.5, side = 4, at = 66)
 ```
 
-Additionally, @fig-plot-ca-loi illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data].
+Interestingly, @fig-plot-loi-map (top) illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data].
 It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis.
@@
 
-The Guttman effect arises from the unimodal distribution of observed variables along a gradient, suggesting the possibility of ordering these variables along the same gradient.
-Our results strongly suggest that the gradient under consideration is related to the firing process: the first axis represents the opposition between minerals present in the raw material and those formed during firing.
-These secondary minerals gradually form during the firing process as conditions change and are likely to be consumed in subsequent chemical reactions.
-Hence, a unimodal distribution can be expected for various $2\theta$ positions along this gradient.
+Therefore, it appears that the ceramics from Mas de Moreno can be ordered according to the progress of these mineralogical transformations or, equivalently, according to the energy received during the firing (@fig-plot-loi-map, bottom).
 
-# DISCUSSION
+## Archaeological implications
+
+The material from the Mas de Moreno is the result of waste from the production process.
+It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect).
+In order to explore these two hypotheses, the diffratograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals.
+This was made possible by the technical homogeneity of the material studied.
+Samples from El Palao and Torre Cremada have the same compositional ranges and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
+
+These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (@fig-plot-guttman D).
+This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; @fig-plot-guttman A and D, @fig-plot-loi-map bottom).
+Individuals reflecting a typical production pattern exhibit peaks related to gehlenite rather than anorthite and diopside.
+This suggests a certain level of control over the energy input during firing to allow the development of this intermediate phase.
+
+The firing process must be carefully controlled to maintain a favorable balance between benefits and risks throughout production.
+In the context of Iberian ceramics, the management of thermal parameters is primarily understood as a strategy to utilize calcareous materials without the risk of lime blowing [@petitdominguez2003, @cultrone2011].
+However, the thermal decomposition of calcium carbonates is a necessary process to achieve the desired characteristics, especially aesthetic qualities.
+
+By incorporating iron, the formation of calcium silicates during firing limits the development of coloration induced by the presence of iron oxides in the matrix.
+This occurs because iron can be integrated into amorphous or crystalline phases through substitution, rather than remaining as free oxides, thereby affecting its chromogenic properties.
+The cations Fe^2+^ and Fe^3+^ are incorporated into pyroxenes (preferably in octahedral coordination) through solid-state diffusion, replacing Mg^2+^ and Ca^2+^ ions, or into melilite-group phases (in tetrahedral coordination).
+This process prevents the formation of spinel compounds [@vedder1969; @nodari2007] and plays a fundamental role in the manufacturing process of Iberian ceramics, as it allows for the attainment of a characteristic pink-colored surface.
+
+Finally, it is worth noting that there appears to be no relationship with chronology (@fig-plot-guttman B).
+Despite significant workshop modifications, including the adoption of a large amphora kiln that resulted in a shift in production quantities during the final years of the workshop's operation [@gorgues2012], the produced artefacts exhibit consistent mineralogical characteristics.
 
@@
 
+The initial application of correspondence analysis to X-ray powder diffraction (XRPD) data of ceramic materials revealed the significant potential of this statistical method in the technological analysis of ancient pottery.
+The main advantage is the possibility of ordering a large number of samples according to a firing gradient.
+This allows for the tracking of mineralogical transformations occurring during firing and facilitates more precise comparisons than those based on estimating the equivalent firing temperature.
+Furthermore, this method enables the construction of calibration data sets from workshops, which can then be utilized to assess the firing of materials discovered in specific contexts.
+
+However, a number of limitations need to be highlighted.
+Firstly, this method only allows us to obtain a relative order, without being able to identify any gaps in the sequence.
+Secondly, this approach relies on homogeneity conditions and mathematical assumptions that can be derived from the conditions formulated by @dunnell1970 for chronological seriation.
+The homogeneity conditions state that all the groups included in such a study must: (1) be of comparable raw materials, (2) belong to the same technological tradition, (3) come from the same local area.
+The mathematical assumptions state that the distribution of any mineralogical class is continuous through firing and exhibits the form of a unimodal curve.
+This last point is undoubtedly the most delicate, as many diffraction angles are paired with several minerals.
+Moreover, diffraction patterns depend not only on the mineralogical composition of the samples, but also on instrumental settings and counting statistics, which currently prevents the reuse of data acquired under different conditions.
+
+These constraints will serve as a guide for future work.
+One of the next steps will be to work on variable selection.
+The use of whole diffractograms leads to the construction of data matrices which tend to be sparse (a large portion of entries consists of zeros) and there is much redundant information.
+Finally, the use of resampling methods, such as bootstrapped approaches, should improve the stability of relative orderings, as shown by @peeples2012.
 # ACKNOWLEDGEMENTS
@@
 
-Code for data cleaning and analysis is provided as compendium in @frerebeau2023a.
+Code for data cleaning and analysis is provided as a compendium in @frerebeau2023a.
 

### 2023-06-08 analysis/supplementary-materials/01-data.R
Add WD-XRF data

@@ -8,6 +8,50 @@ get_bdx <- function(file) {
   regmatches(x = file, m = regexpr(pattern = "BDX[0-9]{5}", text = file))
 }
 
+#' Geometric Mean
+#'
+#' @param x A [`numeric`] vector.
+#' @param trim A length-one [`numeric`] vector specifying the fraction (0 to 0.5)
+#'  of observations to be trimmed from each end of `x` before the mean is
+#'  computed.
+#' @param na.rm A [`logical`] scalar: should `NA` values be stripped before the
+#'  computation proceeds?
+#' @return A [`numeric`] vector.
+#' @noRd
+gmean <- function(x, trim = 0, na.rm = FALSE) {
+  index <- is.finite(x) & x > 0
+  exp(mean(log(unclass(x)[index]), trim = trim, na.rm = na.rm))
+}
+
+# READ WD-XRF ==================================================================
+## List all files
+here::here("analysis/data/raw_data") |>
+  list.files(pattern = "WDXRF", full.names = TRUE) |>
+  lapply(
+    FUN = function(x) {
+      xrf <- read.table(x, header = TRUE, sep = ",", dec = ".")
+      xrf$site <- gsub("\\s*\\([^\\)]+\\)", "", xrf$site)
+      xrf[, c("sample", "site", "CaO", "SiO2", "Al2O3")]
+    }
+  ) |>
+  do.call(rbind, args = _) |>
+  (function(x) split(x, f = x$sample))() |>
+  lapply(
+    FUN = function(x) {
+      if (nrow(x) == 1) return(x)
+      x$sample = unique(x$sample)
+      x$site = unique(x$site)
+      x$CaO = gmean(x$CaO)
+      x$SiO2 = gmean(x$SiO2)
+      x$Al2O3 = gmean(x$Al2O3)
+    }
+  ) |>
+  do.call(rbind, args = _) |>
+  utils::write.table(
+    file = here::here("analysis/data/derived_data/xrf_cas.csv"),
+    sep = ",", dec = ".", row.names = FALSE,
+  )
+
 # READ XRD =====================================================================
 ## List all diffractograms (Brucker RAW)
 xrd_files <- here::here("analysis/data/raw_data") |>
@@ -94,7 +138,7 @@ for (i in seq_along(xrd_data)) {
   selected$y <- selected$y - baseline$y
   lines(selected, col = "blue")
 
-  ## Subset from 5° to 55° (2theta)
[…]

### 2023-06-08 analysis/paper/paper.qmd
Small edits

@@
 bibliography: [references.bib, packages.bib]
-csl: "../templates/elsevier-harvard.csl"
+csl: "../templates/archeosciences.csl"
 abstract: |
-  Text of abstract
+  Abstract
 keywords: |
@@
 |------------:|-----------:|:---------------------------|:--------|
-| 0.0.0.9000  |  |  |  |
+| 0.0.0.9000  |  | Submitted to _Archéosciences, revue d'Archéométrie_ |  |
 
@@
 Quantitative analysis of XRD data faces significant difficulties due to the multiphase nature, poor crystallization, and amorphous content of archaeological ceramics.
-Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material (and may require access to expensive databases).
+Estimating quantitative mineralogical compositions through methods like the RIR method [@hubbard1988] or Rietveld refinement [@rietveld1969] is a challenging and highly skilled exercise, requiring in-depth knowledge of the material.
 
-Limited work has been done on the use of multivariate methods for the analysis of ray diffraction data from archaeological materials.
+Limited work has been done on the use of multivariate methods for the analysis of X-ray diffraction data from archaeological materials.
 Different approaches have been used such as Lineal Discriminant Analysis [@garciagimenez2006], Principal Component Analysis [@delavilla2003; @holakooei2014] or hierarchical clustering [@piovesan2013; @maritan2015], mainly for classification purposes.
@@
 Calibration data sets can be produced solely based on archaeological evidence to assess ceramic firing practices, eliminating the need for laboratory references.
+To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop (Teruel, Aragon, Spain).
 
-To illustrate this method, we examine XRD data of Iron Age ceramics from the Mas de Moreno workshop (Teruel, Aragon, Spain).
-The workshop was active between 225-200 BC and 40-30 AD.
-Previous archaeological studies of the workshop have highlighted three successive phases of activity, characterized by changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012].
-Several kilns have been excavated, with discontinuation and destruction occurring around 50-40 BC [@gorgues2012].
-The use of a larger amphora kiln, the production of Roman amphorae, and the presence of Latin epigraphy indicate a major reorganization of the workshop in the middle of the 1st century.
-The workshop ceased its activity in the following decade due to territorial reorganization under Roman influence and its distance from trade routes [@gorgues2012].
-This workshop provides a context well-suited for investigating the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.
-
-# METHODS
+# MATERIAL and METHODS
 ## Archaeological samples
@@
 
-`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon) were sampled to represent the entire period of the workshop's activity [stratigraphic and typological attributions can be accessed in @frerebeau2015g; @frerebeau2015e].
+`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon) were sampled to represent the entire period of the workshop's activity (@tbl-corpus).
+The workshop was active between 225-200 BC and 40-30 AD.
+Previous archaeological studies of the workshop have highlighted three successive phases of activity, characterized by changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012].
+Several kilns have been excavated, with discontinuation and destruction occurring around 50-40 BC [@gorgues2012].
+The use of a larger amphora kiln, the production of Roman amphorae, and the presence of Latin epigraphy indicate a major reorganization of the workshop in the middle of the 1st century.
+The workshop ceased its activity in the following decade due to territorial reorganization under Roman influence and its distance from trade routes [@gorgues2012].
+This workshop provides a context well-suited for investigating the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.
+
 The ceramic material of the Mas de Moreno consists only of products known as Iberian painted/fine wares.
 It refers to a wheel thrown production, with a pink uniform colour and a dark red painted decor, with no temper visible to the naked eye and considered to be fired at high temperatures in an oxidizing atmosphere [@tarradell1980; @mata1992; @collconesa2000].
+Although Iberian fine wares present a significant technological uniformity [@tarradell1980], the sampling was conducted in such a way that all types of pottery produced in the workshop were included in the study (non-Iberian types produced at the end of the workshop's activity period were also sampled).
+
+A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017].
+Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (@tbl-corpus).
+
+Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (@tbl-corpus).
+These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC).
+Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study.
+Samples from El Palao and Torre Cremada have the same compositional ranges in the CaO-Al~2~O~3~-SiO~2~ system and belong to the same category of Iberian fine wares as the material from Mas de Moreno (see below).
 
@@
 SU: stratigraphic unit.
-LOI: loss on ignition [see @frerebeau2020a]."
-#| eval: false
+LOI: loss on ignition [see @frerebeau2020a; @frerebeau2015e; @frerebeau2015g]."
 corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration", "part"))] |> 
@@
 ```
-
-Although Iberian fine wares present a significant technological uniformity [@tarradell1980], the sampling was conducted in such a way that all types of pottery produced in the workshop were included in the study (non-Iberian types produced at the end of the workshop's activity period were also sampled).
-
-A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017].
-Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here.
-
-Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled.
-These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC).
-Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study.
 
@@
 ## Read mineral data
-minerals <- here::here("analysis/data/derived_data/minerals.csv") |>
+minerals <- here::here("analysis/data/raw_data/minerals.csv") |>
   read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8") |> 
@@
 
-The data were collected at room temperature using a D8 Advance (Bruker) diffractometer in Bragg-Brentano configuration operating at 1.6 kW (40 kV, 40 mA) and equipped with a copper anode source ($k_{\alpha1}$ = 1.5406 Å; the $k_\beta$ ray was removed by a Ni-filter in the diffracted beam).
+The data were collected using a D8 Advance (Bruker) diffractometer in Bragg-Brentano configuration operating at 1.6 kW (40 kV, 40 mA) and equipped with a copper anode source ($k_{\alpha1}$ = 1.5406 Å; the $k_\beta$ ray was removed by a Ni-filter in the diffracted beam).
 An 8 mm anti-scattering slit was mounted in front of the LynxEye© CCD detector.
@@
 
-CA is an effective method for the seriation of archaeological assemblages (REF).
+CA is an effective method for the chronological seriation of archaeological assemblages [see @ihm2005 for an historical overview].
 The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation.
@@
 
-# RESULTS
+# RESULTS AND DISCUSSION
+## Correspondence analysis
 
@@
 The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca A and B).
-Plotting the coordinates of the individuals shows an interesting pattern (@fig-plot-ca C and D).
+Plotting the coordinates of the individuals shows an interesting pattern (@fig-plot-ca C).
 The upper quadrants each consist of a small group of closely located individuals, likely exhibiting high similarity to one another.
@@
 (C) CA rows score plot.
-(D) CA columns score plot."
+(D) CA columns score plot, displaying only the main diffraction lines of a selected set of minerals of interest for improved readability.
+Ilt: illite; Cal: calcite; Qz: quartz; Gh: gehlenite; Di: diopside; An: anorthite."
 #| fig-width: 10
@@
 ## Active rows (F1 vs F2)
-xlim <- range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1], xrd_moreno_col_coord[, 1])
-ylim <- range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2], xrd_moreno_col_coord[, 2])
+
+theta_minerals <- vapply(
+  X = minerals$theta,
+  FUN = function(x, theta) { which.min(abs(theta - x)) },
+  FUN.VALUE = numeric(1),
+  theta = theta_scale
+)
+xrd_minerals_coord <- xrd_moreno_col_coord[theta_minerals, ]
+xlim <- range(range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1]), range(xrd_minerals_coord[, 1]))
+ylim <- range(range(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2]), range(xrd_minerals_coord[, 2]))
+
 plot(
@@
 ## Columns (F1 vs F2)
-plot(
-  x = xrd_moreno_col_coord[, 1],
-  y = xrd_moreno_col_coord[, 2],
-  xlim = xlim, ylim = ylim,
-  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
-  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
-  type = "p",
-  pch = 16,
-  las = 1, asp = 1,
-  col = adjustcolor("black", alpha = 0.2)
-)
-text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
-abline(h = 0, lty = 2, col = "darkgrey")
-abline(v = 0, lty = 2, col = "darkgrey")
-```
-
-@fig-plot-contrib shows the contribution of the different columns to the construction of the first two CA-axes.
-It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix.
-Consequently, the columns that contribute the most to the construction of the first axis (with a contribution greater than 0.5) are associated with the following diffraction peaks: 21.94° (plagioclases), 26.64° (quartz), 27.76° (plagioclases), 27.90° (plagioclases), and 29.42° (calcite).
-Similarly, for the second axis, the columns that contribute the most are those corresponding to the peaks at 12.32° (phyllosilicate), 26.66° (quartz), and 31.36° (melilites).
-
-```{r}
-#| label: fig-plot-contrib
-#| fig-cap: Columns contribution to the construction of the first two components. The twenty variables contributing the most to each component are highlighted in red. 
-#| fig-width: 10
-#| fig-height: 5
-par(mar = c(0, 4, 1, 1), mfrow = c(2, 1), oma = c(3, 0, 0, 0))
-plot(
-  x = theta_scale,
-  y = xrd_moreno_col_contrib[, 1],
-  type = "l",
-  las = 1,
-  ylab = "Contribution to F1",
-  xaxt = "n"
-)
-abline( #21.94, 27.76-27.9, 29.42
-  v = theta_scale[order(xrd_moreno_col_contrib[, 1], decreasing = TRUE)[1:20]],
-  col = adjustcolor(col = "red", alpha = 0.1)
-)
-# pks_contrib1 <- alkahest::peaks_find(
-#   x = theta_scale,
-#   y = xrd_moreno_col_contrib[, 1],
-#   SNR = 2,
-#   m = 5
+# plot(
+#   x = xrd_moreno_col_coord[, 1],
+#   y = xrd_moreno_col_coord[, 2],
+#   xlim = xlim, ylim = ylim,
+#   xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+#   ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+#   type = "p",
+#   pch = 16,
+#   las = 1, asp = 1,
+#   col = adjustcolor("black", alpha = 0.2)
 # )
-# dput(pks_contrib1$x[pks_contrib1$y > 0.5])
-# points(x = pks_contrib1$x[pks_contrib1$y > 0.5],
-#        y = pks_contrib1$y[pks_contrib1$y > 0.5])
-
-plot(
-  x = theta_scale,
-  y = xrd_moreno_col_contrib[, 2],
-  type = "l",
-  las = 1,
-  ylab = "Contribution to F2"
-)
-abline( #12.32, 26.66, 31.36
-  v = theta_scale[order(xrd_moreno_col_contrib[, 2], decreasing = TRUE)[1:20]],
-  col = adjustcolor(col = "red", alpha = 0.1)
-)
-mtext(expression(2*theta), side = 1, line = 2, outer = TRUE)
-# pks_contrib2 <- alkahest::peaks_find(
-#   x = theta_scale,
-#   y = xrd_moreno_col_contrib[, 2],
-#   SNR = 2,
-#   m = 5
-# )
-# dput(pks_contrib2$x[pks_contrib2$y > 0.5])
-# points(x = pks_contrib2$x[pks_contrib2$y > 0.5],
-#        y = pks_contrib2$y[pks_contrib2$y > 0.5])
-```
-
-The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman).
-This effect is characterized by a strong nonlinear relationship between the second factor and the first factor.
-The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014].
-Interestingly, the samples located in the upper right quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A and B, @fig-plot-xrd).
-These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1.
-Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.
-
-```{r}
-#| label: fig-plot-guttman
-#| fig-cap: "Guttman effect observed on the first plane of correspondence analysis.
-Row coordinates are highlighted based on material (A), chronological phase (C), and supplementary individuals with 95% probability ellipses (D).
-In the top right (B), the column coordinates are shown, displaying only the main diffraction lines of a selected set of minerals of interest for improved readability.
-Ilt: illite; Cal: calcite; Qz: quartz; Gh: gehlenite; Di: diopside; An: anorthite."
-#| fig-width: 10
-#| fig-height: 10
-layout(matrix(data = c(1, 3, 2, 4), ncol = 2))
-par(mar = c(4, 4, 1, 1))
-
-theta_minerals <- vapply(
-  X = minerals$theta,
-  FUN = function(x, theta) { which.min(abs(theta - x)) },
-  FUN.VALUE = numeric(1),
-  theta = theta_scale
-)
-xrd_minerals_coord <- xrd_moreno_col_coord[theta_minerals, ]
-xlim <- range(range(xrd_moreno_row_coord[, 1]), range(xrd_minerals_coord[, 1]))
-ylim <- range(range(xrd_moreno_row_coord[, 2]), range(xrd_minerals_coord[, 2]))
-
-id_moreno <- match(rownames(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]), corpus$sample)
-
-## F1 vs F2: active only, by material
-col_material <- khroma::color("high contrast")(2)
-plot(
-  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
-  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
-  xlim = xlim, ylim = ylim,
-  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
-  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
-  type = "p",
-  pch = c(16, 17)[as.factor(corpus$material[id_moreno])],
-  las = 1, asp = 1,
-  col = col_material[as.factor(corpus$material[id_moreno])]
-)
-abline(h = 0, lty = 2, col = "darkgrey")
-abline(v = 0, lty = 2, col = "darkgrey")
-text(x = min(xlim), y = max(ylim) + 0.25, labels = "A", font = 2, cex = 1.5)
-legend("bottomright", legend = levels(as.factor(corpus$material[id_moreno])),
-       pch = c(16, 17), col = col_material)
-
-## Columns (F1 vs F2)
-par(mar = c(4, 4, 1, 1))
+# text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
+# abline(h = 0, lty = 2, col = "darkgrey")
+# abline(v = 0, lty = 2, col = "darkgrey")
 col_minerals <- khroma::color("bright")(6)
@@
 text(x = min(xlim), y = max(ylim) + 0.25, labels = "B", font = 2, cex = 1.5)
+```
+
+@fig-plot-contrib shows the contribution of the different columns to the construction of the first two CA-axes.
+It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix.
+Consequently, the columns that contribute the most to the construction of the first axis (with a contribution greater than 0.5) are associated with the following diffraction peaks: 21.94° (plagioclases), 26.64° (quartz), 27.76° (plagioclases), 27.90° (plagioclases), and 29.42° (calcite).
+Similarly, for the second axis, the columns that contribute the most are those corresponding to the peaks at 12.32° (phyllosilicate), 26.66° (quartz), and 31.36° (melilites).
+
+```{r}
+#| label: fig-plot-contrib
+#| fig-cap: Columns contribution to the construction of the first two components. The twenty variables contributing the most to each component are highlighted in red. 
+#| fig-width: 10
+#| fig-height: 5
+par(mar = c(0, 4, 1, 1), mfrow = c(2, 1), oma = c(3, 0, 0, 0))
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 1],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F1",
+  xaxt = "n"
+)
+abline( #21.94, 27.76-27.9, 29.42
+  v = theta_scale[order(xrd_moreno_col_contrib[, 1], decreasing = TRUE)[1:20]],
+  col = adjustcolor(col = "red", alpha = 0.5)
+)
+# pks_contrib1 <- alkahest::peaks_find(
+#   x = theta_scale,
+#   y = xrd_moreno_col_contrib[, 1],
+#   SNR = 2,
+#   m = 5
+# )
+# dput(pks_contrib1$x[pks_contrib1$y > 0.5])
+# points(x = pks_contrib1$x[pks_contrib1$y > 0.5],
+#        y = pks_contrib1$y[pks_contrib1$y > 0.5])
+
+plot(
+  x = theta_scale,
+  y = xrd_moreno_col_contrib[, 2],
+  type = "l",
+  las = 1,
+  ylab = "Contribution to F2"
+)
+abline( #12.32, 26.66, 31.36
+  v = theta_scale[order(xrd_moreno_col_contrib[, 2], decreasing = TRUE)[1:20]],
+  col = adjustcolor(col = "red", alpha = 0.5)
+)
+mtext(expression(2*theta), side = 1, line = 2, outer = TRUE)
+# pks_contrib2 <- alkahest::peaks_find(
+#   x = theta_scale,
+#   y = xrd_moreno_col_contrib[, 2],
+#   SNR = 2,
+#   m = 5
+# )
+# dput(pks_contrib2$x[pks_contrib2$y > 0.5])
+# points(x = pks_contrib2$x[pks_contrib2$y > 0.5],
+#        y = pks_contrib2$y[pks_contrib2$y > 0.5])
+```
+
+The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman).
+This effect is characterized by a strong nonlinear relationship between the second factor and the first factor.
+The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014].
+Interestingly, the samples located in the upper right quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd).
+These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1.
+Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.
+
+```{r}
+#| label: fig-plot-guttman
+#| fig-cap: "Guttman effect observed on the first plane of correspondence analysis: row coordinates are highlighted based on material (A), chronological phase (B), and supplementary individuals with 95% probability ellipses (C).
+CaO-SiO2-Al2O3 ternary plot of the ceramic samples (D).
+Qz: quartz; Gh: gehlenite; An: anorthite; Wo: wollastonite; mul: mullite; crn: corundum.
+Data from @frerebeau2015g and @frerebeau2015e."
+#| fig-width: 10
+#| fig-height: 10
+layout(matrix(data = c(1, 3, 2, 4), ncol = 2))
+par(mar = c(4, 4, 1, 1))
+
+xlim <- range(range(xrd_moreno_row_coord[, 1]))
+ylim <- range(range(xrd_moreno_row_coord[, 2]))
+
+id_moreno <- match(rownames(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]), corpus$sample)
+
+## F1 vs F2: active only, by material
+col_material <- khroma::color("high contrast")(2)
+plot(
+  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
+  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
+  xlim = xlim, ylim = ylim,
+  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
+  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
+  type = "p",
+  pch = c(16, 17)[as.factor(corpus$material[id_moreno])],
+  las = 1, asp = 1,
+  col = col_material[as.factor(corpus$material[id_moreno])]
+)
+abline(h = 0, lty = 2, col = "darkgrey")
+abline(v = 0, lty = 2, col = "darkgrey")
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "A", font = 2, cex = 1.5)
+legend("bottomright", legend = levels(as.factor(corpus$material[id_moreno])),
+       pch = c(16, 17), col = col_material)
 
@@
 abline(v = 0, lty = 2, col = "darkgrey")
-text(x = min(xlim), y = max(ylim) + 0.25, labels = "C", font = 2, cex = 1.5)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "B", font = 2, cex = 1.5)
 legend("bottomright", 
@@
 abline(v = 0, lty = 2, col = "darkgrey")
-text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "C", font = 2, cex = 1.5)
 legend("bottomright", legend = c(levels(as.factor(corpus$site[corpus$site != "Mas de Moreno"])), "Mas de Moreno"),
@@
 
+## Ternary plot
+xrf_cas <- here::here("analysis/data/derived_data/xrf_cas.csv") |> 
+  read.table(header = TRUE, sep = ",", dec = ".") |> 
+  subset(sample %in% corpus$sample)
+
+par(mar = c(1, 1, 1, 1))
+isopleuros::ternary_plot(
+  xrf_cas[xrf_cas$site == "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
+  col = "lightgrey", pch = 16, ylab = ""
+)
+isopleuros::ternary_points(
+  xrf_cas[xrf_cas$site != "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
+  pch = c(24, 25)[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])],
+  bg = col_site[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])]
+)
+isopleuros::triangle_phase_ceramic(symbol = TRUE)
+text(x = 0, y = 0.9, labels = "D", font = 2, cex = 1.5)
 ```
@@
 
-# DISCUSSION
 ## Ceramic firing
@@
 The principal axis of correspondence analysis is built upon the opposition between primary phases and secondary phases.
-The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (@fig-plot-guttman B).
+The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (@fig-plot-ca D).
 The intermediate position of gehlenite can also be explained.
-The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle [@frerebeau2020a].
+The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991 [@frerebeau2020a]: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle (@fig-plot-guttman D).
 In this case, the formed gehlenite is metastable.
@@
 This was made possible by the technical homogeneity of the material studied.
-Samples from El Palao and Torre Cremada have the same compositional ranges and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
+Samples from El Palao and Torre Cremada have the same compositional ranges (@fig-plot-guttman D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
 
-These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (@fig-plot-guttman D).
-This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; @fig-plot-guttman A and D, @fig-plot-loi-map bottom).
+These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (@fig-plot-guttman C).
+This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; @fig-plot-guttman A, @fig-plot-ca D, @fig-plot-loi-map bottom).
 Individuals reflecting a typical production pattern exhibit peaks related to gehlenite rather than anorthite and diopside.
@@
 The firing process must be carefully controlled to maintain a favorable balance between benefits and risks throughout production.
-In the context of Iberian ceramics, the management of thermal parameters is primarily understood as a strategy to utilize calcareous materials without the risk of lime blowing [@petitdominguez2003, @cultrone2011].
+In the context of Iberian ceramics, the management of thermal parameters is primarily understood as a strategy to utilize calcareous materials without the risk of lime blowing [@petitdominguez2003; @cultrone2011; @cultrone2014].
 However, the thermal decomposition of calcium carbonates is a necessary process to achieve the desired characteristics, especially aesthetic qualities.
@@
 
-The initial application of correspondence analysis to X-ray powder diffraction (XRPD) data of ceramic materials revealed the significant potential of this statistical method in the technological analysis of ancient pottery.
+The initial application of correspondence analysis to X-ray powder diffraction data of ceramic materials revealed the significant potential of this statistical method in the technological analysis of ancient pottery.
 The main advantage is the possibility of ordering a large number of samples according to a firing gradient.
@@
 Secondly, this approach relies on homogeneity conditions and mathematical assumptions that can be derived from the conditions formulated by @dunnell1970 for chronological seriation.
-The homogeneity conditions state that all the groups included in such a study must: (1) be of comparable raw materials, (2) belong to the same technological tradition, (3) come from the same local area.
+The homogeneity conditions state that all the groups included in such a study must: (1) be of comparable raw materials, (2) belong to the same technological tradition, (3) show mineralogical transformations related to firing.
 The mathematical assumptions state that the distribution of any mineralogical class is continuous through firing and exhibits the form of a unimodal curve.
@@
 One of the next steps will be to work on variable selection.
-The use of whole diffractograms leads to the construction of data matrices which tend to be sparse (a large portion of entries consists of zeros) and there is much redundant information.
+The use of whole diffractograms leads to the construction of a data matrix which tend to be sparse (a large portion of entries consists of zeros) and there is much redundant information.
 Finally, the use of resampling methods, such as bootstrapped approaches, should improve the stability of relative orderings, as shown by @peeples2012.
+
 # ACKNOWLEDGEMENTS
@@
 
-The data that support the findings of this study are openly available in [@frerebeau2015h] and [@frerebeau2015i].
+The data that support the findings of this study are openly available in @frerebeau2015g, @frerebeau2015e, @frerebeau2015h, @frerebeau2015i.
 
-Code for data cleaning and analysis is provided as a compendium in @frerebeau2023a.
+The R code for data cleaning and analysis is openly available as a compendium in @frerebeau2023a. It allows to replicate all the results presented above.
 

### 2023-06-08 README.Rmd
Use renv

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
-
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Frerebeau, N. (2023). _`r Title`_. _Archéosciences, revue d'Archéométrie_. <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Frerebeau, N. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-## How to run in your browser or download and run locally
+## How to download and run locally
 
@@
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+[main.zip](/archive/main.zip). After unzipping:
 - open the `.Rproj` file in RStudio
@@
 - finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+
+To be as portable and reproducible as possible, this compendium uses the [`renv`](https://rstudio.github.io/renv/) package to manage the packages used.
+Opening the project with RStudio should trigger the installation of `renv`, the necessary packages can then be installed with :
+
+``` r
+renv::restore()
+```
 

### 2023-06-08 analysis/paper/paper.qmd
Add abstract

@@
 abstract: |
-  Abstract
+  The study of the firing of archaeological ceramics is thus a key factor in 
+  interpreting ancient techniques and related social practices. This study 
+  proposes a new method based on the correspondence analysis of powder X-ray 
+  diffraction data (XRD), aiming to provide an alternative to the so-called
+  Equivalent Firing Temperature framework. The properties of correspondence 
+  analysis and its application for matrix seriation enable the description of 
+  the firing of a ceramic assemblage. The application of correspondence 
+  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
+  of this statistical method in the technological analysis of ancient pottery. 
+  This allows to order of a large number of samples according to a firing 
+  gradient and to track mineralogical transformations occurring during firing.
 keywords: |
   X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
-highlights: |
-  These are the highlights. 
 ---
@@
 Keywords: `r rmarkdown::metadata$keywords`
-
-```{r}
-#| label: graphical-abstract
-#| fig-align: center
-#| out.width: "4in"
-knitr::include_graphics(here::here("analysis/figures/abstract.png"))
-```
 
@@
 |------------:|-----------:|:---------------------------|:--------|
-| 0.0.0.9000  |  | Submitted to _Archéosciences, revue d'Archéométrie_ |  |
+| 0.0.0.9000  |  | Submitted to _Archéosciences_ |  |
+
+```{r}
+#| label: graphical-abstract
+#| fig-align: center
+#| out.width: "4in"
+knitr::include_graphics(here::here("analysis/figures/abstract.png"))
+```
 
@@
 
-# INTRODUCTION
+# Introduction
 
@@
 
-# MATERIAL and METHODS
+# Material and method
 ## Archaeological samples
@@
 
-# RESULTS AND DISCUSSION
+# Results and discussion
 ## Correspondence analysis
@@
 
-# CONCLUSION
+# Conclusion
 
@@
 
-# ACKNOWLEDGEMENTS
+# Acknowledgements
 
@@
 
-# DATA AVAILABILITY STATEMENT
+# Data availability statement
 
@@
 
-# REFERENCES
+# References
 

### 2023-06-08 analysis/paper/paper.qmd
Add figures

@@
     reference-doc: "../templates/template.docx" # Insert path for the DOCX file
-execute:
+knitr:
+  opts_chunk:
     echo: false
@@
     comment: "#>"
-  fig-path: "../figures/"
-  fig-dpi: 600
+    fig.path: "../figures/"
+    fig.dpi: 600
 filters:

### 2023-06-09 analysis/paper/paper.qmd
Small edit

--- 2023-06-08 analysis/paper/paper.qmd
@@
 abstract: |
-  The study of the firing of archaeological ceramics is thus a key factor in 
+  The study of the firing of archaeological ceramics is a key factor in 
   interpreting ancient techniques and related social practices. This study 
@@
   gradient and to track mineralogical transformations occurring during firing.
+  This method should let to more precise comparisons than those based on 
+  estimating the equivalent firing temperature.
 keywords: |
@@
   read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8")
+
+## Read XRF data
+xrf_cas <- here::here("analysis/data/derived_data/xrf_cas.csv") |> 
+  read.table(header = TRUE, sep = ",", dec = ".") |> 
+  subset(sample %in% corpus$sample)
 ```
@@
 #| label: tbl-corpus
-#| tbl-cap: "Ceramic and clay samples analyzed in this study.
+#| tbl-cap: "Ceramic and clay (unfired pottery sherds) samples analyzed in this study.
 Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC.
-SU: stratigraphic unit.
-LOI: loss on ignition [see @frerebeau2020a; @frerebeau2015e; @frerebeau2015g]."
-corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration", "part"))] |> 
+LOI: loss on ignition [@frerebeau2020a; @frerebeau2015g; @frerebeau2015e]."
+corpus |> 
+  merge(y = xrf_cas[, -2], by = "sample") |> 
+  (function(x) x[order(x$sample), ])() |> 
+  (function(x) x[, !(colnames(x) %in% c("site_type", "stratigraphy", "artefact", "decoration", "part"))])() |> 
+  (function(x) {
+    write.table(x, file = here::here("analysis/tables/corpus.csv"), 
+                col.names = TRUE, row.names = FALSE, sep = ",", dec = ".")
+    x
+  })() |> 
   knitr::kable(
-    col.names = c("Sample", "Material", "Site", "SU", "Phase", "Type", "LOI (%)"),
+    col.names = c("Sample", "Material", "Site", "Phase", "LOI (%)", "CaO (%)", "SiO2 (%)", "Al2O3 (%)"),
     booktabs = TRUE,
@@
 ## Ternary plot
-xrf_cas <- here::here("analysis/data/derived_data/xrf_cas.csv") |> 
-  read.table(header = TRUE, sep = ",", dec = ".") |> 
-  subset(sample %in% corpus$sample)
-
 par(mar = c(1, 1, 1, 1))
@@
 
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
-::: {#refs}
-:::
-
-\newpage
-
 # Colophon {-}
@@
 ```
+
+\newpage
+
+# References

### 2023-06-09 analysis/supplementary-materials/01-data.R

@@ -23,36 +23,8 @@ gmean <- function(x, trim = 0, na.rm = FALSE) {
   exp(mean(log(unclass(x)[index]), trim = trim, na.rm = na.rm))
 }
 
-# READ WD-XRF ==================================================================
-## List all files
-here::here("analysis/data/raw_data") |>
-  list.files(pattern = "WDXRF", full.names = TRUE) |>
-  lapply(
-    FUN = function(x) {
-      xrf <- read.table(x, header = TRUE, sep = ",", dec = ".")
-      xrf$site <- gsub("\\s*\\([^\\)]+\\)", "", xrf$site)
-      xrf[, c("sample", "site", "CaO", "SiO2", "Al2O3")]
-    }
-  ) |>
-  do.call(rbind, args = _) |>
-  (function(x) split(x, f = x$sample))() |>
-  lapply(
-    FUN = function(x) {
-      if (nrow(x) == 1) return(x)
-      x$sample = unique(x$sample)
-      x$site = unique(x$site)
-      x$CaO = gmean(x$CaO)
-      x$SiO2 = gmean(x$SiO2)
-      x$Al2O3 = gmean(x$Al2O3)
-    }
-  ) |>
-  do.call(rbind, args = _) |>
-  utils::write.table(
-    file = here::here("analysis/data/derived_data/xrf_cas.csv"),
-    sep = ",", dec = ".", row.names = FALSE,
-  )
-
-# READ XRD =====================================================================
+# XRD ==========================================================================
+## READ ------------------------------------------------------------------------
 ## List all diffractograms (Brucker RAW)
 xrd_files <- here::here("analysis/data/raw_data") |>
   list.files(pattern = ".raw", full.names = TRUE)
@@ -89,7 +61,7 @@ for (i in xrd_data) {
   lines(x = i[[1]], y = i[[2]])
 }
 
-# PRE-PROCESS ==================================================================
+## PRE-PROCESS -----------------------------------------------------------------
 ## Penalized likelihood smoothing
 lambda <- seq(from = 1, to = 8, length.out = 40)
 lambda <- 10^lambda
@@ -156,7 +128,7 @@ for (i in seq_along(xrd_data)) {
   dev.off()
 }
 
-# WRITE ========================================================================
+## WRITE -----------------------------------------------------------------------
[…]

### 2023-06-09 analysis/paper/paper.qmd
Prepare for first release

@@
 |------------:|-----------:|:---------------------------|:--------|
-| 0.0.0.9000  |  | Submitted to _Archéosciences_ |  |
+| 1.0.0       | 2023-06-09 | Submitted to _Archéosciences_ |  |
 

### 2023-06-09 README.Rmd
Prepare for first release
Add DOI

@@
 
-> Frerebeau, N. (2023). _`r Title`_. _Archéosciences, revue d'Archéométrie_. <https://doi.org/xxx/xxx>
+> Frerebeau, N. (2023). _`r Title`_. _Archéosciences, revue d'Archéométrie_.
 
@@
 
-> Frerebeau, N. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Frerebeau, N. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/10.5281/zenodo.8020137>
 
@@
 **Data :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
-
-### Contributions
-
-We welcome contributions from everyone.
-Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
-Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md).
-By participating in this project you agree to abide by its terms.

### 2023-06-09 analysis/paper/paper.qmd
Add DOI

--- 2023-06-09 analysis/paper/paper.qmd
@@
 | **Version** | **Date**   | **Description**            | **DOI** |
-|------------:|-----------:|:---------------------------|:--------|
-| 0.0.0.9000  |  | Submitted to _Archéosciences_ |  |
+|---------:|-----------:|:------------------------------|:---------------------------|
+| 1.0.0    | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020137](https://doi.org/10.5281/zenodo.8020137) |
 

### 2023-08-01 analysis/paper/paper.qmd
Edits after reviews.

--- 2023-06-09 analysis/paper/paper.qmd
@@
   of this statistical method in the technological analysis of ancient pottery. 
-  This allows to order of a large number of samples according to a firing 
+  This allows to order a large number of samples according to a firing 
   gradient and to track mineralogical transformations occurring during firing.
@@
 | **Version** | **Date**   | **Description**            | **DOI** |
-|------------:|-----------:|:---------------------------|:--------|
-| 0.0.0.9000  |  | Submitted to _Archéosciences_ |  |
+|---------:|-----------:|:------------------------------|:---------------------------|
+| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020137](https://doi.org/10.5281/zenodo.8020137) |
+| 1.1      | 2023-08-01 | Corrections based on reviews  | |
 
@@
 rows) in the same way.
-Unlike PCA, CA analyzes the differences between relative values and considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space).
-CA studies the inertia (i.e., the weighted sum of the χ2 distances of each point to the centroid) to create orthogonal components so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the second component, and so on until the last dimension.
+Unlike PCA, CA analyzes the differences between relative values and considers both individuals and variables at the same time.
+This allows for the projection of row and column points in the same coordinate space, so that the relative positions of one set of points provide the reading keys for interpreting the positions of the other set of points along the axes [@greenacre2007].
+
+CA studies the inertia (i.e., the weighted sum of the $\chi^2$ distances of each point to the centroid) to create orthogonal components so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the second component, and so on until the last dimension.
 CA maximizes the correspondence between individuals and variables instead of maximizing the amount of variance explained by its reduced space.
@@
 The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time.
-CA is not limited to chronological modeling but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e., any aspect that is expected to be related to the observed composition (be it temporal or environmental).
+CA is not limited to chronological modelling but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e., any aspect that is expected to be related to the observed composition (be it temporal or environmental).
 
@@
 The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca.
-Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and shards.
+Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and sherds.
 The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals.
@@
 )
-text(x = min(xlim), y = max(ylim) + 0.25, labels = "B", font = 2, cex = 1.5)
+text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
 ```
@@
 The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014].
-Interestingly, the samples located in the upper right quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd).
+Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd).
 These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1.
@@
 It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect).
-In order to explore these two hypotheses, the diffratograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals.
+In order to explore these two hypotheses, the diffractograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals.
 This was made possible by the technical homogeneity of the material studied.

### 2023-08-01 analysis/paper/paper.qmd
Update changelog

@@
 |---------:|-----------:|:------------------------------|:---------------------------|
-| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020137](https://doi.org/10.5281/zenodo.8020137) |
-| 1.1      | 2023-08-01 | Corrections based on reviews  | |
+| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020138](https://doi.org/10.5281/zenodo.8020138) |
+| 1.1      | 2023-08-01 | Corrections based on reviews  | [10.5281/zenodo.8204409](https://doi.org/10.5281/zenodo.8204409) |
 

### 2023-12-04 analysis/paper/paper.Rmd
Use Rmarkdown instead of Quarto

--- 2023-08-01 analysis/paper/paper.qmd
@@
       orcid: 0000-0001-5759-4944
-      affiliations:
-        - name: "UMR 6034 Archéosciences Bordeaux"
-          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
-          city: Pessac cedex
-          country: France
-          postal-code: 33607
-          url: https://www.archeosciences-bordeaux.fr
-title-block-published: "Last updated"  
-date: now
-date-format: long
-format: 
-  pdf: 
-    documentclass: scrartcl
-    include-in-header: 
-      - text: |
-          \usepackage{booktabs}
-    classoption: 
-      - french
-      - 12pt
-      - a4paper
-      - oneside
-  docx:
-    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
-knitr:
-  opts_chunk:
-    echo: false
-    warning: false
-    message: false
-    comment: "#>"
-    fig.path: "../figures/"
-    fig.dpi: 600
-filters:
-  - ../templates/scholarly-metadata.lua
-  - ../templates/author-info-blocks.lua
-  - ../templates/pagebreak.lua
+      institute:
+        - AB
+institute:
+  - AB:
+      name: UMR 6034 Archéosciences Bordeaux
+      address: Maison de l'Archéologie, Université Bordeaux Montaigne, 33607 Pessac cedex, France
+date: "Accepted for publication on 2023-09-15"
 bibliography: [references.bib, packages.bib]
@@
   X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
+output: 
+  bookdown::pdf_document2:
+    toc: false
+    fig_caption: yes
+    number_sections: true
+    pandoc_args:
+    - --lua-filter=../templates/scholarly-metadata.lua
+    - --lua-filter=../templates/author-info-blocks.lua
+    - --lua-filter=../templates/pagebreak.lua
 ---
@@
 #| label: setup
+#| include: false
 options(knitr.kable.NA = "")
+
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  collapse = FALSE,
+  comment = "#>",
+  fig.path = "../figures/",
+  fig.dpi = 600
+)
 ```
@@
 Keywords: `r rmarkdown::metadata$keywords`
+
+```{r}
+#| label: graphical-abstract
+#| fig-align: center
+#| out.width: "4in"
+knitr::include_graphics(here::here("analysis/figures/abstract.png"))
+```
 
@@
 |---------:|-----------:|:------------------------------|:---------------------------|
-| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020137](https://doi.org/10.5281/zenodo.8020137) |
-| 1.1      | 2023-08-01 | Corrections based on reviews  | |
-
-```{r}
-#| label: graphical-abstract
-#| fig-align: center
-#| out.width: "4in"
-knitr::include_graphics(here::here("analysis/figures/abstract.png"))
-```
+| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020138](https://doi.org/10.5281/zenodo.8020138) |
+| 1.1      | 2023-08-01 | Corrections based on reviews  | [10.5281/zenodo.8204409](https://doi.org/10.5281/zenodo.8204409) |
+| 1.2      | 2023-09-15 | Accepted for publication      |  |
 
@@
 
-`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon) were sampled to represent the entire period of the workshop's activity (@tbl-corpus).
+`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon; table \@ref(tab:tbl-corpus)) were sampled to represent the entire period of the workshop's activity [@frerebeau2020a; @frerebeau2015g].
 The workshop was active between 225-200 BC and 40-30 AD.
@@
 
-A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017].
-Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (@tbl-corpus).
+A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017; @frerebeau2015e].
+Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (table \@ref(tab:tbl-corpus)).
 
-Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (@tbl-corpus).
+Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (table \@ref(tab:tbl-corpus)).
 These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC).
@@
 #| label: tbl-corpus
-#| tbl-cap: "Ceramic and clay (unfired pottery sherds) samples analyzed in this study.
-Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC.
-LOI: loss on ignition [@frerebeau2020a; @frerebeau2015g; @frerebeau2015e]."
 corpus |> 
@@
     col.names = c("Sample", "Material", "Site", "Phase", "LOI (%)", "CaO (%)", "SiO2 (%)", "Al2O3 (%)"),
+    caption = "Ceramic and clay (unfired pottery sherds) samples analyzed in this study.
+Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC.
+LOI: loss on ignition.",
     booktabs = TRUE,
@@
 
-The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca.
+The results of the correspondence analysis of the diffractograms are presented in fig.
+\@ref(fig:fig-plot-ca).
 Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and sherds.
@@
 
-The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca A and B).
-Plotting the coordinates of the individuals shows an interesting pattern (@fig-plot-ca C).
+The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (fig.
+\@ref(fig:fig-plot-ca)A and B).
+Plotting the coordinates of the individuals shows an interesting pattern (fig.
+\@ref(fig:fig-plot-ca)C).
 The upper quadrants each consist of a small group of closely located individuals, likely exhibiting high similarity to one another.
@@
 
-@fig-plot-contrib shows the contribution of the different columns to the construction of the first two CA-axes.
+fig.
+\@ref(fig:fig-plot-contrib) shows the contribution of the different columns to the construction of the first two CA-axes.
 It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix.
@@
 
-The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman).
+The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (fig.
+\@ref(fig:fig-plot-ca), \@ref(fig:fig-plot-guttman)).
 This effect is characterized by a strong nonlinear relationship between the second factor and the first factor.
 The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014].
-Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd).
+Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (fig.
+\@ref(fig:fig-plot-guttman)A, \@ref(fig:fig-plot-ca)D, \@ref(fig:fig-plot-xrd)).
 These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1.
@@
 #| label: fig-plot-xrd
-#| fig-cap: "Diffractograms of representative samples from each quadrant of the CA plot (see @fig-plot-ca).
+#| fig-cap: "Diffractograms of representative samples from each quadrant of the CA plot (see fig.
+\\@ref(fig:fig-plot-ca)).
 From top to bottom: upper left quadrant (BDX14922), lower left quadrant (BDX15451), lower right quadrant (BDX15465), upper right quadrant (BDX22203).
-Only the peaks that contribute the most to the construction of the first two CA axes are indexed (see text and @fig-plot-contrib)."
+Only the peaks that contribute the most to the construction of the first two CA axes are indexed (see text and fig.
+\\@ref(fig:fig-plot-contrib))."
 #| fig-width: 10
@@
 The principal axis of correspondence analysis is built upon the opposition between primary phases and secondary phases.
-The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (@fig-plot-ca D).
+The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (fig.
+\@ref(fig:fig-plot-ca)D).
 The intermediate position of gehlenite can also be explained.
-The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991 [@frerebeau2020a]: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle (@fig-plot-guttman D).
+The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991 [@frerebeau2020a]: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle (fig.
+\@ref(fig:fig-plot-guttman)D).
 In this case, the formed gehlenite is metastable.
@@
 
-Interestingly, @fig-plot-loi-map (top) illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data].
+Interestingly, figure \@ref(fig:fig-plot-loi-map) (top) illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data].
 It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis.
@@
 
-Therefore, it appears that the ceramics from Mas de Moreno can be ordered according to the progress of these mineralogical transformations or, equivalently, according to the energy received during the firing (@fig-plot-loi-map, bottom).
+Therefore, it appears that the ceramics from Mas de Moreno can be ordered according to the progress of these mineralogical transformations or, equivalently, according to the energy received during the firing (fig.
+\@ref(fig:fig-plot-loi-map), bottom).
 
@@
 This was made possible by the technical homogeneity of the material studied.
-Samples from El Palao and Torre Cremada have the same compositional ranges (@fig-plot-guttman D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
+Samples from El Palao and Torre Cremada have the same compositional ranges (fig.
+\@ref(fig:fig-plot-guttman)D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
 
-These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (@fig-plot-guttman C).
-This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; @fig-plot-guttman A, @fig-plot-ca D, @fig-plot-loi-map bottom).
+These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (fig.
+\@ref(fig:fig-plot-guttman)C).
+This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; fig.
+\@ref(fig:fig-plot-guttman)A, \@ref(fig:fig-plot-ca)D, \@ref(fig:fig-plot-loi-map) bottom).
 Individuals reflecting a typical production pattern exhibit peaks related to gehlenite rather than anorthite and diopside.
@@
 
-Finally, it is worth noting that there appears to be no relationship with chronology (@fig-plot-guttman B).
+Finally, it is worth noting that there appears to be no relationship with chronology (\@ref(fig:fig-plot-guttman)B).
 Despite significant workshop modifications, including the adoption of a large amphora kiln that resulted in a shift in production quantities during the final years of the workshop's operation [@gorgues2012], the produced artefacts exhibit consistent mineralogical characteristics.
